\documentclass[11pt]{article}
\usepackage{amssymb} % has to be used before XeTeX unicode trickery.
\usepackage{fullpage}
\usepackage{fontspec}
\usepackage{xunicode}
\usepackage{xltxtra}
\defaultfontfeatures{Mapping=tex-text}
%\setromanfont{Linux Libertine O}
\setromanfont{Brill}
%\newfontfamily\greek{Gentium}


\usepackage{multicol}
\usepackage{ulem}
\usepackage{ifthen}
% Interdocument linking.
\usepackage[xetex]{hyperref}
\hypersetup{bookmarksopen=false,
  pdfpagemode=UseNone,
  colorlinks=true,
  urlcolor=black,
  linkcolor=black,    % no links for footnotes; URLs will still have color
  pdftitle={Hypmnemata Glossopoetica},
  pdfauthor={William S. Annis},
  pdfkeywords={conlang}%,
}


% Better sectioning for this document.
\usepackage[compact,rigidchapters,explicit]{titlesec}
\setcounter{secnumdepth}{4}
\titleformat{\section}[display]
 {\normalfont\fillast}
 {\normalfont\bfseries \thesection. #1}
 {1ex minus .1ex}
 {\small}
\titlespacing{\section}{3pc}{*4}{-1em}[3pc]

\titleformat{\subsection}[runin]{\normalfont\bfseries}{\thesubsection.}{.5em}{#1. }[ { }]
\titlespacing{\subsection}{1ex}{1.5ex plus .1ex minus .2ex}{0pt}

\titleformat{\subsubsection}[runin]{\normalfont\bfseries\small}{\thesubsubsection.}{.5em}{#1. }[ { }]
\titlespacing{\subsubsection}{1ex}{1.5ex plus .1ex minus .2ex}{0pt}

% If no argument is given, only the section is printed, no title.
\titleformat{\paragraph}[runin]{\normalfont\bfseries\small}{\theparagraph.}{.5em}{\ifthenelse{\equal{#1}{}}{}{#1. }}[ { }]
\titlespacing{\paragraph}{1ex}{1.5ex plus .1ex minus .2ex}{0pt}

% Some utilities.
\newcommand{\LL}[1]{\textbf{#1}}  % Other language
\newcommand{\E}[1]{\textit{#1}}   % English
\newcommand{\I}[1]{\textsc{#1}}   % Interlinears
\newcommand{\note}[1]{\textcolor{magenta}{\small\textit{#1}}}
\newcommand{\tsref}[1]{\hyperref[#1]{\S \textbf{\ref*{#1}}}}
\newcommand{\interlin}[1]{\begin{quotation}{\small\noindent#1}\end{quotation}}
\newcommand{\rara}[1]{$\mathfrak{R}$: #1}
\newenvironment{grammarlist}%
 {\begin{itemize}\addtolength{\itemsep}{-0.5\baselineskip}\ignorespaces}%
 {\end{itemize}\ignorespacesafterend}

\newenvironment{dlist}%
 {\begin{quote}\begin{description}\addtolength{\itemsep}{-0.3\baselineskip}\ignorespaces}%
 {\end{description}\ignorespacesafterend\end{quote}}

\newenvironment{examples}{\quote}{\endquote}
\newcommand{\example}[2]{\noindent\LL{#1}\hskip1em\E{#2}}
\newcommand{\longexample}[2]{\noindent\LL{#1}
\indent\E{#2}}

\begin{document}
\frenchspacing
\title{Hypomnemata Glossopoetica}
\author{Wm S. Annis}
\date{\today}
\maketitle

\section{Phonology}

Normally illegal clusters may occur in particular grammatical
contexts, and thus look common (\textit{cf.} Latin \LL{-st} in \I{3sg}
copula \LL{est}).

% http://www.unish.org/upload/word/사본%20-%20MarkVanDam%5B1%5D.pdf
Hierarchy of codas:\footnote{Some attested single-\I{c} coda
  inventories: \{n ŋ\}, \{n ŋ t k\}, \{n ŋ m p t k\}, \{n m l r\}, \{n
  m w j\}, \{n ŋ m l r j\}, \{n m r d\}, \{d l s x\}, \{m b k l z r\},
  \{n l w j t k\}, \{n ŋ m l p t k\}, \{w n m r k t v ʃ ʒ\}.} n < m,
ɳ, ŋ < ɳ << l, ɹ < r, ʎ, ʁ < ɭ, ɽ << t < k, p < s, z, c, q, ʃ < b, d,
g, x h << w, j.  There is a slight place hierarchy: alveolar < velar <
retroflex or tap.  Classes percolate, such that in complex codas, if
Nasals, Resonants and Stops are permitted you usually expect \I{n, r,
  s, nr, ns} and \I{rs} as coda sequences.  Other orders are possible,
but the above rule is common-ish.

Hierarchy of clusters (\I{s} = sonorant, \I{o} = obstruent), word
initial: \I{os} < \I{oo} < \I{ss} < \I{so}; word final: \I{so} <
\I{oo} < \I{ss} < \I{os}.  Onset clusters tend to avoid identical
places of articulation, which leads to avoidance of things like
\textit{*tl, dl, bw,} etc., in a good number of languages. /j/ is
lightly disfavored as \I{c2} after dentals, alveolars and palatals;
/j/ and palatals are in general disfavored before front vowels.

Languages with \textit{sC-} clusters often have codas.  s+\I{stop} <
s+\I{fric} / s+\I{nasal} < s+\I{lat} < s+\I{rhot} (the fricative and
nasal are trickier to order).

Even if a particular \I{c} is a permitted coda, its allowed
environment may be quite restricted. Potential constraints: forbidden
before homorganic stop; or homorganic nasal; geminates
forbidden. Solutions: delete with compensatory vowel lengthening;
debuccalize (become fricative, glottal stop, delete without
compensation); nasal deletion with nasalized vowel remaining;
tone wackiness.

Lower vowels are preferred as syllabic nuclei; high vowels are more
prone to syncope (either midword or finally).  Content words less
likely to elide.
% https://linguistics.stonybrook.edu/sites/default/files/uploads/u26/publications/QP1FinalDraft.pdf

Sonority hierarchy:
\begin{quotation}
\noindent low vowels > mid vowels (except /ə/) > high vowels (except /ɨ/) > \\
\indent /ə/ > /ɨ/ > glides > laterals > \\
\indent flaps > trills > nasals > /h/ > \\
\indent voiced fricatives > \\
\indent voiced stops and affricates, voiceless fricatives > \\
\indent voiceless fricatives, voiced stops and affricates > \\
\indent voiceless stops and affricates
\end{quotation}

Vowel devoicing starts with lowest sonority  (/i/, /u/).

% https://www.ling.upenn.edu/~gene/courses/530/readings/Casali2011.pdf
Hiatus resolution: 1) elide, contract, add glide /j/ or /w/, with the
glide sharing the same frontness or roundness of \I{v1}; 2)
insert /h/ or /ʔ/; 3) insert coronal consonant, usually /t/ or some
/r/.

Metathesis (where P = /p, b, m/, etc.).  Obstruent + resonant can
switch.  \I{pk} > \I{kp} occasionally, but not the other way.
\I{t\{pk\}} > \I{\{pk\}t} reasonably common.

% http://linguistics.berkeley.edu/phonlab/annual_report/documents/2007/Hyman_Phono_Universals_PL.pdf
Many things can happen to a consonant following a nasal: NT > ND; NS,
NZ > NTS, NDZ; NT > NTʰ; ND > NN; but also: ND > NT; NTʰ > NT; NN >
ND.

Fortition is much less common that lenition.  Most commonly
strengthened is approximant → fricative: (/j/ → [ʑ zʲ $^\textrm{d}$j
dʒ]; /w/ → [β v ᶢw]).  Change most common before high and/or front
vowels, or at onset (word or syllable).  Fricative strengthening less
common: /ʃ/ → [tʃ]; /s/ -> [dʒ] (before sonorant); /ɸ/ → [p] word
initial, or before back vowels.
% https://doi.org/10.1515/lingty-2019-0015 - Consonant strengthening:
% A crosslinguistic survey and articulatory proposal 

Consonant strength hierarchy (in some African languages, generalized):
continuants, semivowels and ʔ > voiced stops > nasals > voiced and
nasal geminates, voiceless stops > voiceless geminates > nasal
clusters.\footnote{Bilin has consonant ``ablaut'' in the plural, where
  the penultimate or ultimate consonant (occasionally both) altnernate
  in the plural: \textit{b-f, d-t, d-s, d-ʃ, r-t, l-t, r-l, dʒ-ʃ, g-k,
    x-k, gʷ-kʷ, xʷ-kʷ, x-k', xʷ-k'ʷ, w-kʷ}.}
% http://daberinet.com/Blin-files/Main-p-files/Blin%20Morphology%20,%20by%20David%20L.%20Appleyard.pdf

Some suffixes may occur only in pausa, or are different there.

Syllable weight: closed syllables may be heavy or light depending on
the coda consonant, with sonorants more often making weight and stops
not (in general CVV > CVC > CV).  Also, low vowels are generally
heavier than high, and central vowels less heavy than non-central. You
can get word-length dependent affix allomorphs of the same number of
syllables, but with heavier vowels in one and lighter in the other.

Consider palatalization, labialiazation, nasalization, etc., as
phenomena which apply to a \textit{syllable} rather than a particular
segment. This can generate substantially differing outcomes in
historical processes (\textit{v.} Chadic family), especially as
differing branches apply the effects more closely or more distantly. 
% https://www.eva.mpg.de/lingua/conference/08_springschool/pdf/course_materials/Wolff_Historical_Phonology.pdf
% "Issues in the historical phonology of Chadic languages"

For determining stress accent, Malagasy has some light final
syllables. These are the remnants of final consonants that got vowels
tacked on after a switch to a dominant \I{cv} structure (and the other
final consonants deleted).

\subsection{ATR Harmony}
Some systems are dominated by the feature of the word root, causing
harmony both to the left and the right. Other systems are based on
syllable position, \textit{e.g.,} a system where \I{-atr} vowels can
follow, but not precede, \I{+atr} vowels. An originally larger vowel
inventory may reduce, giving a situation where one apparent vowel can
trigger \I{+atr} in some words and \I{-atr} in others. Different
morphological systems may take different rules within one language.

May apply to all vowels, or just to \I{-atr} /ɪ ʊ/ (or /ɛ ɔ/).

/a/ may be transparent, or may block. Or may be \I{-atr}, with /ə/ the
\I{+atr} version.

Common systems: 9V /iɪeɛaɔoʊu/; 7V(M) /ieɛa(ə)ɔou/ or 7V(H)
/iɪɛa(ə)ɔʊu/. \I{-atr} usually dominant in 7V(M), \I{+atr} usually
dominant in 7V(H).

\subsection{Tone}
Tone inventory hierarchy: \I{rf, fr > r > f > h, l} (\I{r}ising,
\I{f}alling).

In many Dogon languages, postnominal adjectives and demonstratives
(but not numbers or other quantifiers other than one) cause all tones
to lower in the preceding \I{np}.  If multiple adjectives follow, all
are lowered except the last. The head noun of an internally headed
\I{rc} is also so lowered.
% http://www.dartmouth.edu/%7Emcpherson/892heath.pdf - "Tonosyntax and
% reference restriction in Dogon NPs".

\subsection{Reduplication} The original word-initial consonant of a
reduplicated root may undergo lenition, or fortition.  Malagasy
\LL{fantatra} > \LL{\uline{f}anta\uline{p}antatra}.

The \textbf{inflectional} uses include: number (for nouns; for verbs
marking actor number or event plurality; in noun compounds either or
both elements may reduplicate, sometimes in free variation); very
rarely to encode possession (either of 3rd person, or 1st and 2nd,
usually with partial reduplication); frequent in verbs for
frequentative, habitual or progressive aspect, but also imperfective,
inchoative and perfect(ive); Margi participles are formed by partial
or full \I{red} (seen in a few other languages).  In Tabasaran
(Lezgian) some verbs of 3 syllables reduplicate the second for
negation in some non-finite forms.

\I{red} may be used to repair a word shape that has taken some affix.

% http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.54.8895&rep=rep1&type=pdf
The \textbf{derivational} uses include: ordinal numbers from cardinal;
distribution (``3 each''); valency reduction in verbs (including
antipassive), as well as reciprocals (and mutuality in other word
classes, ``face to face''); diminutive (and both endearment and
contempt); associatives (``someone with X''); similarity; sort/kind of
N; disposition (``someone prone to X, likely to X''); common in
indefinites (``who'' > ``whoever''); in verbs, a lack of control,
disorder, carelessness, pretense, attempt; incrementality
(``gradually, little by little, one by one''); spread-out or scattered
(``here and there, looking around''); non-uniformity (``zig-zag, now
and then, in several colors, hodge-podge'').

Not infrequently used to name insects and birds, and the word for
``baby.'' 

Noun to verb (``to wear an X''), adjective or adverb; verb to noun
(agent noun, action noun, instrument), adjective or adverb; adjective
to adverb.

For adjectives and other vocabulary of quality: diminutive
attentuative, augmentative, intensification.

The sense of ``collectivity'' might be used where we would expect a
single term (``broom, flight of stairs,'' etc).

Some affixes may simply require the root take \I{red}, with Makah
having several types of \I{red} and affixes selecting one.


\section{Word Classes}

\begin{center}
  \small
  \begin{tabular}{|l|l|l|l|}
    \hline
    & \LL{Reference} & \LL{Modification} & \LL{Predication} \\
    \hline
\LL{Object} & referring, argument phrase (\LL{noun}) & nominal attributive phrase &
                                                                   predicate
                                                                   nominal \\
    \hline
\LL{Property} & deadjectival nominal & attributive phrase (\LL{adjective}) & predicate adjectival\\
    \hline
\LL{Action} & complement clause & relative clause & clause (\LL{verb})\\
    \hline
  \end{tabular}
\end{center}


\LL{Universal One}. A lexical class used in a nonprototypical
propositional act function\footnote{Reference, modification or
predication.} will be coded with at least as many morphemes as in
its prototypical function, as in \textit{bright $>$ bright-\LL{ness}}
(prototypical functions are marked in bold in the chart).

\LL{Universal Two}. A lexical class used in a nonprototypical
propositional act function will also have no more grammatical
behavioral potential than in its prototypical function. For example, a
predicate nominal will not have more verbal inflectional possibilities
than a full verb.

Individual languages will break up that 9x9 grid differently in terms
of constructions, and where individual words end up.


\section{The Noun}
Nouns are the most frequently borrowed word class.

\subsection{Possession}
``Possession'' can cover a wide range of relationships:

\begin{grammarlist}
  \item Ownership (or temporary possession)
  \item Whole-part relationship (body part, part of an object)
  \item Kinship
  \item Attribute of a person, animal or thing (``Bob's temper'')
  \item Orientation or location (``the front of the house''; useful
    when body part terms are used for location to use different
    marking)
  \item Association (``my teacher,'' but also dwellings, house to
    homeland, and personal clothing and goods)
  \item Nominalization
\end{grammarlist}

\noindent Most languages do not have the wide range of possessive uses
found in English or Greek in the same construction.  The first three
in the list above are most central.

Marking for inalienable possession is generally smaller (fewer
syllables, no classifier, simpler construction) than the marking for
alienable.

Genitive marked on R, pertensive marked on D.  Affixed possession
markers may also induce pertensive marking.  Genitive marking
frequently has other functions, pertensive rarely.

Different systems of marking may occur in two or even three groups
depending on possessor: pronoun, proper noun, kin term, common human
noun, common animate noun, common inanimate noun (with contiguous
constructions between groups).

Splits across whole-part D: external body parts, internal body parts,
genitalia, body fluids, parts of animals, parts of plants, parts of
artifacts. 

\subsection{Classifiers}
Number classifiers are most common with ``one'' and ``two,'' possibly
obligatory only with these; rarely with different classifier forms for
higher numbers.  More classifier types may be available with lower
numbers.

With stative verbs, the classic Dixon set are most likely to take
classifiers; or postural verbs.  With transitive verbs, more likely on
high-agency, high-patientness verbs (handling, cooking, killing and
the like).  With any verb class, optional classifiers may mark
salience or ``completeness'' of the classified feature.

When not used with their expected word class, classifiers may be
derivational.

\subsubsection{Gender / Noun Class}
Variable gender assignment may code size and/or shape and/or posture
(upright vs.\ horizontal).  Phonological gender assignment (initial,
final, or both, are possible determinates) typically restricted to
nonanimates.  Function of referent may change gender (water as drink
vs.\ water as part of the landscape).

Agreement: clefting may interfere or inhibit agreement; mixed class
resolution may default to the least marked form, but the behavior for
animate vs.\ inanimate may pick different forms, and coordination of
mixed animacy may simply be avoided altogether (a different
construction, or repeating the phrase with different subjects, etc.);
phonetics may inhibit some kinds of agreement.  Semantic agreement
override: attributive < predicate < relative < personal pronoun.

Different word classes may pick different genders for the least
marked, catch-all form.

It is rare but possible to have two noun class systems operating in a
single language, with different agreement rules operating on different
word classes.  In particular, ``pronominal'' agreement for pronouns
and verb agreement\footnote{Paumarí has m./f.\ for demonstratives, but
a separate \textit{ka}/non-\textit{ka} agreement system on the
verbs, showing up as a prefix.} and ``nominal'' agreement for
adjectives (and sometimes numbers).  Pronominal agreement in general
is focused on pronouns, is a smaller system, aligns with animacy, sex
or humanness, and has a fairly transparent semantic basis.  Nominal
agreement systems tend to be larger, focus on animacy, sex, shape and
size, and the semantics may be much less clear.  Demonstratives align
with either pattern, from language to language.  Where the two systems
have semantic overlap, the markers may be quite different, or cover
different spaces.

Hierarchy (tendency) for site of marking: pronouns > verbs >
demonstratives > adjectives > numerals. Also possible: verbs >
adjectives > pronouns.

\subsubsection{Noun Classifiers}
Similar to gender, and historically may lead to it, but classifiers
are different.  The semantics are generally clearer; some languages
may allow cooccurance of classifiers (\I{specific general}, as in
``\I{human man} boy''); single \I{n} may take different classifiers to
determine precise meaning; may be used anaphorically (esp.\ across
clauses of a single sentence), possibly with other syntactic functions
(relative clauses).  Attend to coreferentiality.

May be small (2--3) or larger (20, or more).

``Social status'' may be encoded in human classifiers (``initiate,
known,'' etc.).  Some classifiers encode inherent nature (person, bug,
tree), some function (edible, drinkable, movable, etc.).  Multiple
classifiers will usually encode one of each.

Classifier on \I{n} may be omitted once class established, or
obligatory always.

May be separate \I{q} for ``unknown class'' vs.\ ``known class but no
more.'' 

Noun classifier systems often related to number classifier systems,
but they may be different.  Number classifiers are much less likely to
be optional.  Both systems may occur in a single \I{np}.

\subsubsection{Number Classifiers}
Sometimes there is a catch-all classifier, sometimes many nouns don't
take any classifier.

May not go beyond ten, or are dropped with 10s units.

Classifiers for humans/animates may have different forms with
different numbers; with low numbers the \I{n + cl} form can be
suppletive.

Affixed classifiers more likely to code animacy, and to be required;
independent classifiers code shape, consistency, etc., and may be more
or less optional, with different speakers at different competencies.
Rarely, affixed and independent classifiers may appear together.

A single noun may take different classifiers to focus salient
characteristic.

\subsubsection{Possessive Classifiers}
Can code: shape, consistence, animacy of possessum; relationship of
possessor to possessum (as in Oceanic indirect possession); or, very
rarely, class of possessor.

There may just be a bunch of words for ``of'' which match class
(possibly highly suppletive).

\subsubsection{Verbal Classifiers} 
Many originate from noun incorporation, possibly competing with it.
However, some verbal classifiers may be clearly related to numeral or
other classifiers in the language.  Another origin is from verbs, via
grammaticalized serial verb constructions.

Existential verbs may distinguish animate from inanimate; a few
languages elaborate existentials quite a lot (in a container, movable
vs.\ immobile, non-human animal, etc.).


\subsection{Case}
\I{sov} languages more likely to have case marking than \I{svo}.

\subsubsection{Formal Marking}
\begin{center}
  \begin{tabular}{ccccl}
    A & O & E & peripheral \\
    \hline
    w & x & y & z & Latin (very common) \\
    w & x & \uline{y} & \uline{y} & Jarawara \\
    w & \uline{x} & \uline{x} & y & Kinyarwanda \\
    w & \uline{x} & \uline{x} & \uline{x} & Creek
  \end{tabular}
\end{center}
% http://www.smg.surrey.ac.uk/features/morphosemantic/transitivity/

\noindent In Creek there is \LL{-t} on subjects and \LL{-n} on
nonsubjects. 

\subsubsection{Accusative} This may be sensitive to the animacy of the
object, with high animacy objects more likely to take an overt
accusative (so-called ``differential object marking'').  \I{dom} may
be sensitive to the relative animacy of \I{a} and \I{o}.  Or it may
vary with definiteness.  Even in languages with \I{dom}, \I{o} marking
may be required in all ditransitive constructions.

In many languages, \I{acc} may be used on adverbs of distance and
duration.  It is also regular on nominalized complement clauses (``I
know he-is-here-\I{acc}'').

Sometimes motion verbs take a goal in the \I{acc}.

Historically, \I{acc} may evolve from \I{dat}, more rearely from
\I{inst}. 

% http://www.ling.helsinki.fi/~eniirane/sija/37-Spencer-ch36.pdf
% https://www.researchgate.net/publication/297883853_Case_Polysemy

\subsubsection{Allative} A huge range of possibilities here.
Occurring more often: true allative (``to, towards, reaching for''),
purpose (``use it for that, in order to''), conceptual (``think about,
occur to''), recipient (dative), timepoint (``at \I{time}''),
addressee (``talk to me''), perceptual (``look at, listen to''),
reason (``because of, ran from fear'').  Some less frequent
possibilities include: temporal boundary (``by/until \I{time}''),
benefactive, possessive, porportion or rate (``3 out of 4, 3 at a
time''), equivalence (``as, in exchange for''), subordinator
(``although, when, while''), emotional target (``hard for, be angry
at'').
% RiceKabata2007.pdf

\subsubsection{Ablative} An ablative may mark not only the source but
also the path of motion (perlative).

\subsubsection{Locative} Very rarely (Páez) a language may have
several locatives which mark posture (standing, lying, hanging,
leaning, etc.).
% http://www.deniscreissels.fr/public/Creissels-spatial_cases.pdf

\subsubsection{DPM} Differental Place Marking is possible, with the
hierarchy: human > common inanimate > place.  Shorter locative marking
becomes more likely as you go down the hierarchy (such as Latin using
bare cases without prepositions for the names of towns and small
islands, and a few things like \textit{domus}).  Names of cities seem
to come in for this a bit more than other things.  Further, there
might be special additional morphology when animates take locative
marking, such as the Basque \textit{-ga(n)-} extension for \I{loc},
\I{abl}, and \I{all}.  Different, often more complex, adpositional
constructions might be required for animate place targets.
% https://dlc.hypotheses.org/2385
% https://www.academia.edu/37762885/Differential_place_marking_and_differential_object_marking

\subsubsection{Vocative} Vocative expressions often have
characteristic vowel lengthening, stress shifts, or tone changes.  In
support of such pitch alterations, overt vocative morphology will
usually have fewer consonants than other sorts of case marking.  Mid
vowels are weakly preferred to high.
% https://psyarxiv.com/ef52k/

Swedish has both long and short forms of ``father'' and ``mother,''
\textit{fader:far, moder:mor}.  The long forms cannot be used as
vocatives (except \textit{fader} in some archaic relgious forms).


\section{The Pronoun}
There may be separate forms for ``\I{prn} alone'' and/or ``\I{prn}
also'' unrelated to similar expressions for nouns.

Languages with 1/2 systems are not common, but not rare either.

Sg/du/paucal/pl is far more common that sg/du/trial/pl.

It's not especially common for sg and pl forms to be related (as in
Chinese).

Many languages:

\begin{center}
\begin{tabular}{lll}
  & Singular & Plural \\
\I{1st} & 1 & 12, 13 (and 11) \\
\I{2nd} & 2 & 22, 23 \\
\I{3rd} & 3 & 33
\end{tabular}
\end{center}

However, one might get something like:

\begin{center}
\begin{tabular}{ll}
1 & 13 (and 11) \\
2 & 12, 22, 23     
\end{tabular}
\end{center}

The basic 1/2/3 sg/pl system may be extended to include a special 12
``me and you'' form.  This may involve the innovation of a dual
throughout the pronoun system.  Or it may be just a normal system with
an inclusive/exclusive distinction.  Or a minimal/augmented system, 

\begin{center}
\begin{tabular}{ll}
Minimal & Non-minimal \\
1  & 1 + others \\
1+2 & 1+2 + others \\
2 & 2 + others \\
3 & 3 + others 
\end{tabular}
\end{center}

\noindent The non-minimal may include ``unit augmented'' (one other
person, producing a dual in many forms) and just ``augmented'' (the
``plural'').  Minimal/augmented are common in Australia, Austronesian,
South America, and a few in North America, though the unit
augmented/augmented distinction is nearly restricted to Australia.

3sg = 3pl is a common neutralization.  2pl = 3pl (and 2du = 3du) can
happen.

Some small pronoun systems (where a number means no plurality
distinction): \I{1ex, 1in, 2, 3; 1sg, 1pl, 2, 3; 1sg, 23sg.m. 23sg.f,
1pl, 23pl.m, 23pl.f}.  The system 1/2/3, with no number distinction,
occurs a few places.  Some of these systems distinguish ``I'' from
``not-I'' only.  However, 12/3 systems also occur.  Quite rich pronoun
systems might still lack a separate 3rd person form (using, say, a
demonstrative).  Independent, citation forms might be lacking
altogether, with nominalized inflected forms of ``be'' used instead
(\I{part-1}-be-\I{nmlz}).
% Parameters of Poor Pronoun Systems, Daniel Harbour.

Using 2du, 2paucal or 2pl as a mark of respect occurs in Europe,
Oceania, Australia.

There is a relationship between the indefinite ``someone'' and
1non-sg forms (French; Caddo, etc).  Usually it's 1pl.inc that does
this. 

Question-based indefinites may be marked with some other morpheme,
often related to words for ``be,'' ``want,'' ``perhaps,'' ``or'' or
``also.''


\subsection{Demonstratives and Deixis}
% https://research.jcu.edu.au/lcrc/storeroom/sashas-folder/demonstratives-directionals-summary ``Demonstratives and directionals: summing up''

\begin{grammarlist}
  \item two way
  \item three way: proximal, distal, remote; proximal, medium, further
    away or imprecise; proximal, medial (not far, known to 1 and 2),
    distal
  \item four way: proximal (to 1), proximal (to 2), medial, distant
  \item five way: priximal visible, proximal audible, medial, distant,
    imperceptible
  \item six way (Godoberi): close to 1, close to 2, that at some
    distance from 1, that at some distance from 2, that down there,
    that (aforementioned)
\end{grammarlist}

The distance required to be excluded from proximal may vary from
within arm's reach (or even in hand) to within view.  Gesture might
decide.  Similarly, different indication gestures might be used with
different distances (lips/head vs.\ fingers, for example).  Possible
to have two proximals, one for things in hand, one for things merely
near.
% https://www.youtube.com/watch?v=rppyxPmwvy0

Verticality expressed: \I{up/down} > \I{level/across}.  Elevation
values usually coexpressed with distal rather than medial or proximal
(that is, elevation is relevant when the referent is not in the
immediate and visible area).  Full mix and match morphology (distance
morphs + elevation morphs) is not at all common, but does occur.
\I{up} demonstratives tend to mean the future, \I{down} the past.  The
\I{up/down} distinction might be general (eye level), or geographic
(up a hill).  Might shrink to micro level (in a single, level house)
based on orientation in the landscape (door facing a river is
\I{down}, say).  Possible to have separate general and geographic
sets.  These are mostly \textit{not} used to indicate social status
(but does occur rarely).
% https://www.frontiersin.org/articles/10.3389/fpsyg.2020.01712/full#T4
% "Elevation as a Grammatical and Semantic Category of
% Demonstratives," Diana Forker.

Rarely, pronominal uses may require nominalizer.

Some systems have forms specifically used for anaphora, or only permit
particular forms to be used anaphorically (slight preference for
distal forms for this?).

In a very few languages, there may be special forms always accompanied
by physical gestures (English ``yei high'').

May have overtones of familiarity, endearment, pejorative,
empathy.  Like diminutives, only context may make clear a positive
vs.\ negative interpretation.  Even proximal deixis can be
pejorative. 

Demonstratives can be recruited as filler, both within a sentence (a
particular form will be preferred), or as ``um, ah.''

Splits: more distinctions for modifier, with fewer for independent,
pronoun-like forms. 

Adverbial locatives (``here, there'') often match demonstrative
distinctions (not infrequently by derivation or clear relation), but
may have fewer distinctions.  Rarely, particular clause positions may
have fewer options (Jarawara, one way distinction clause initial,
``here/there,'' two way distnction clause final).

Manner adverbials (``thus, like this, like that''), again often
related to pronouns, may make fewer distinctions.

Verbal demonstratives: ``do/be (like) this/that.'' Can be both
anaphoric and cataphoric.

\rara{Demonstrative agreement with addressee (in addition to usual),
  with traces of such agreement for \I{adj}.}
% https://orientalberber.wordpress.com/2012/05/29/siwi-addressee-agreement-and-addressing-aljazeera/

\rara{Ik has demonstratives that locate objects in time, with a
five way (non-past, recent past, removed past, remote past, remotest
past) distinction that matches the temporal adverbs.}
% http://langsci-press.org/catalog/book/98


\subsection{Indefinites}
Cross-linguistically, indefinites derived from interrogatives are about
twice as common as those derived from nouns. If a derivational element
is used on interrogative indefinites, often related to ``be, want,
perhaps, or,'' or ``also.''  Kannada uses ``or'' clitic to form
``some'' indefinites, ``also'' clitic for ``any'' indefinites.  A few
languages split function, with interrogatives for ``any'' indefinites
and nouns for ``some'' indefinites.

If a language has \I{q}-derived indefinites, there may still be some
additional marking of a question when used interrogatively, often the
same \I{q} particle used in polar questions.  Or an interrogative
mood.  Also \I{q}-word focus constructions (fronting, particle, clause
intonation) may distinguish.
% https://ling-blogs.bu.edu/lx500f10/files/2010/10/lx500univf10-07-indefinites-handout.pdf

Indefinites off the the left of the Haspelmath map are more likely to
be appreciative, and those to the right depreciative.
% http://www.philol.msu.ru/~otipl/new/fdsl/abstracts/bylinina.pdf



\section{The Adjective}
Comparatives neither rare nor universal.  Can be marked with (1) affix
and adpositional phrase; (2) adposition only; (3) coordination (``X is
big but Y is small'' = ``X is bigger than Y'').

Breaks along word class distinctions follow the adjective class: full
\I{nav} (all of noun, adjective and verb in the lexicon); \I{n[av]}
(i.e., property-concept words go with verbs); \I{[na]v}; and \I{[nav]}
(all three classes conflated).
% http://www.ualberta.ca/~dbeck/StJuste.pdf

Resultative adjectives are most likely in languages in which there are
complex verbal predicates (serial verbs and particle verbs). ``I
hammered it flat'' slots into complex verbal constructions naturally.
French, for example, simply doesn't have these.
% https://www.uni-salzburg.at/fileadmin/multimedia/Linguistik/documents/On_predicting_resultative_adjective_constructions-June-2016.pdf


\section{Numerals}
Bases found in human languages: 2, 3, 4, 5 (hand), 6, 8 (like 4,
counting the spaces between digits), 12, 20, 60.  Hybrids: 5+20(+80),
2+5, 10+20.  Base 10 is by far the most common.  Lower bases are less
common, and will generally only go up to a few powers, with a base 4
system, for example, only reaching 8, 16, or 32.

The higher power units may not be usable alone. That is, unlike
English ``ten,'' some languages may require it be ``one ten'' in the
same way 20 is ``two tens.''

The words for 2 and 2nd often convey sense of ``another''
(``secondary''). 

Overcounting: some systems overshoot, then work backwards, so that
``seven thirty'' is 27 (old Turkic).  Subtraction: 8 = two from ten;
31 = one plus 10 from 2*20.

A very few languages (PNG) use different bases for counting different
things (as in Bukiyip, which uses 3 and 4).

% http://sajms.com/wp-content/uploads/2016/03/A-Typology-of-Rare-features-in-Numerals.pdf ``A typology of rare features in numerals''

\section{The Verb}
``Labile transitivity'' is very common.  Be clear on \I{s = a}
\textit{vs.} \I{s = o} for intransitive constructions.  Some languages
may skew in favor of one or the other, but others do not.  English
``cook,'' with transitive and both intransitives (``the chef cooks;
the chef cooks stew; the stew is cooking'') is unusual.

Lability is a dodge!  If the language has rich mechanisms for changing
transitivity, it's less frequent.

Labile verbs tend to cluster semantically, 1) destruction and strong
property change, ``break, boil, freeze, dry, go out, melt, dissolve,
burn, destroy, break, split, kill/die;'' 2) motion and spatial
configuration, ``rock, roll, sink, spread, close, open, connect,
rise/raise, stop, fill, turn;'' 3) phase, ``begin, finish;'' 4)
non-physical effect, ``change, improve, develop;'' and 5) verbs with
an animate patient, ``wake up, learn, gather.''  A single language may
pick several clusters, or only a few items from one.
% Letuchiy, A., "Interpreting the spontaneity scale."

A transitive verb may be used intransitively in an extended
intransitive sense, too.

Verb pairs (``die/kill,'' ``eat/feed'') for which the S/O role are
primarily animate are more likely to pattern with one basic
transitivity for the simplest stem (i.e., ``kill'' is primary with a
derivation for ``die''), while those with primarily inanimate S/O role
verbs (``boil/boil,'' ``burn/burn'', ``fall/drop'') will also have a
primary transitivity (in English, these are often labile or
suppletive).  In general, though, the basic role for animate verbs
will be intransitive, with augmentation for the transitive form.  On
the other hand, in the large, some few languages are fairly
intransitive, some quite strongly transitive.

Stative verbs often mark fewer tense or aspect distinctions (such as
Eng.\ -ing, ``*I am knowing''). Or, they may take some additional
marking obligatorily (Turkish).

Among other categories, some languages have verbs that mark ease or
difficulty.  The marking for ``ease'' on a transitive verb may signal
a small \I{o}.

In richly marked verbs, these may not be marked: \I{3rd} inanimate
sobjects, all \I{3sg/pl} objects, \I{3sg/pl} topical subjects,
\I{3sg/pl} absolutives, all \I{3sg/pl} of any kind.


\subsection{Affix Order}
The fewer person markers there are, the more likely they are to be
prefixing.  But: object marking on the verb is prefixing with more
than chance frequency.  Prefixing in general is less common with
\I{ov} than with \I{vo}, except for object marking, with a distinct
preference for object prefixing with \I{ov}.
% https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.616.5686&rep=rep1&type=pdf
% http://cysouw.de/home/presentations_files/cysouwASSYMETRY_handout.pdf

If \I{t} and \I{a} are on the same side of the verb, \I{a} is closer;
if different, the order is \I{t Verb a}.

\subsection{Person Marking}
Ways to express the subject of a verb, with overt subject vs.\ no
overt subject:
\begin{grammarlist}
  \item Marta came-she / came-she (70\%)
  \item Marta came-she / she came-she (very rare outside Germanic sphere)
  \item Marta came / she came (7\%ish)
  \item Marta came / came-she (14\%)
  \item Marta came / came (7\%ish)
\end{grammarlist}
% https://dlc.hypotheses.org/1340

Marking of both \I{a} and \I{p} is somewhat more common than all other
options: no agreement, marking \I{a} only, marking \I{p} only (rare),
marking of either (very rare).  Agreement may be either by affixes, or
by clitics that will tend to attach themselves prosodically to the
\I{vp}, or in Wackernagel position.

With objects especially, but subjects also sometimes, the verbal
marking may only occur as a pro-form replacing a missing overt object;
or according to some feature of the arguments, such as animacy,
definiteness, high referential prominence, word class (\I{np}
vs.\ \I{pro}); or by tense, aspect, polarity, mood, or even clause
type.

Grammaticalization path of \I{o} marking is probably different from
that of \I{s/a}.
% https://dlc.hypotheses.org/1304


\subsection{Tense, Aspect, Mood}

\subsubsection{Aspect} Perfective marking has a tense hierarchy: past
> future > present (and the present may be reinterpreted, for example
as habitual or future).

\subsubsection{Telicity}
Tense and default aspect interpretations of verbs may be very
dependent on telicity.
\begin{center}
  \I{(pfv)} Achievement > Accomplishment > Activity > State \I{(ipfv)}
\end{center}
\noindent The most natural location for perfective marking is on
achievement verbs; the most natural for imperfective is state (though
state verbs themselves can be odd).  Cross-marking (\I{ipfv} on
achievement verb) may have other interpretations (such as iterativity
in English ``I was sneezing'').

Im/perfectives may have a default tense interpretation (present/past),
which can split anywhere along the hierarchy above.  In Even, aorist
activities are present, aorist achievements and accomlishments are
recent past.
% http://web.philo.ulg.ac.be/lediasema/wp-content/uploads/sites/43/2018/09/Malchukov_2018_LeDiasema.pdf

\subsubsection{Mood}
It's not unusual in a single language for there to be more kinds of
constructions for expressing mood and the future than for tense and
aspect.

``Irrealis'' often associated with future, or becomes future
historically. 

\paragraph{Imperative}
Implicational map: \I{2sg} > \I{2pl} > \{\I{3sg/pl}, {\I{1pl.incl} >
\I{1pl.excl}\} > \I{1sg}.  The pragmatics of different person
imperatives can be quite distinct, giving ``imperatives'' high
polysemy.  Starting point for each is 1) an ``authorizer'' of an
activity and 2) some agent(s) that can bring about the state of
affairs.

\textbf{\I{2nd} Imperative.} Paradigmatic imperatives. Some
extensions.  1) Hypothetical Imperative: an invitation to imagine a
state of affairs (also possible in \I{3rd}).  This leads to 2)
conditional or concessive imperative, in which the subject is
pragmatically generic, ``give me a place to stand and I will move the
world.'' 2) Rhetorical Imperative: when the action is difficult or
impossible to achieve, with the imperative a (possibly ironic)
invitation to verify this.  Further a mirative imperative is possible
(``just try to find a job these days!'').  3) Echoic Imperative, where
the imperative is used to report a situation where the imperative
might reasonably have been used.  English uses a construction like
``it's a matter of,'' ``they didn't want to understand the principles,
it was a matter of give me the recipe.''  4) Historical or Narrative
Imperative, which is a subset of echoic (Palestinian Arabic, Balkan,
and Romance languages): it may have a sense of obligatory mood,
``everyone is having a rest but I have to work(-\I{imp}).''  Might be
called necessitive, and in Balkans has sense of resentment at
unreasonable demands.

\textbf{\I{3rd} Imperative.} The agentive addressee is expected to
enable some event, rather than handle it directly, or to give
permission.  Extensions: 1) hypothetical (as \I{2nd}), but these are
not inherently generic.  2) mirative ``then suddenly he said(-\I{imp})
to me.''

\textbf{\I{1excl} Imperative.} These are centrally a request for
permission (thus, common markers for \I{3rd} and \I{1excl} forms).
Can be used in deontic (``deliberative'') questions, ``What should I
write(-\I{imp}) in my report?''  Also possible in polar questions.
This deliberative use does not occur in all languages with \I{1excl}
imperatives.  Further extension, is to question the appropriateness of
an action, ``I must watch?'' (I watch(-\I{imp})).  Both deontic and
appropriateness use may involve adverbs or other particles to
distinguish. 

% https://www.researchgate.net/publication/366548658_Towards_an_enhanced_semantic_map_for_imperatives


\paragraph{Frustrative}
A frustrative may be a separate word, a clitic, or (most common) an
inflection.  As usual, a range of meanings possible, with no
expectation that a single language will use a frustrative for all
senses (mostly derived from Amazonian data):

\begin{grammarlist}
  \item action was accomplished, but without expected result (often
    unspecified, but inferrable from context)
  \item action was accomplished but the speaker finds it irritating or
    inconvenient (in a few languages, this sense only in negative clauses)
  \item speaker's unwilling participation
  \item incompletive, the action was started but not accomplished
    (``nearly'') 
  \item incompletive, the expected action didn't even start
  \item counterfactual
\end{grammarlist}
% https://www.academia.edu/28327536/A_typology_of_frustrative_marking_in_Amazonian_languages

If cliticized to a noun, often depreciative rather than simply scoping
over clause.

May be restrictions on combining with: tense, polarity, mood (such as,
never with imperative, or never in future).


\subsection{Evidentiality, Mirativity, Egophoricity}
% http://www.diva-portal.org/smash/get/diva2:821317/FULLTEXT01.pdf
All of evidentiality, mirativity, and egophoricity potentially have
polysemous marking, so that certain uses of, say, evidential marking
might be used for a mirative.  May be merged with rest of TAM system,
or not.

Evidentials may distinguish: visual evidence, non-visual or sensory,
inference, assumption (including reasoning, assumption, or general
knowledge), hearsay, and quotative for overt reference to the source.
Evidentials may be limited to particular tenses or aspects; least
common in the future.  Most common two-way distinction is reported
vs.\ everything else.  Common three-way: direct, hear-say, inference.
Evidentiality is more worried about uncertain information than certain
— if a language has only one evidential, it is most likely to be
indirect. 

Egophoric marking\footnote{Some scholars consider egophore a kind of
  person marking.  This seems less likely as more data comes in.}
prototypically marks \I{ego} for first person and \I{nonego} for
everything else in statements, and marks \I{ego} for second person and
\I{nonego} for everything else in questions.  Evidential distinctions
more likely in \I{nonego}.  Again, past is most likely to mark the
most distinctions.  Might be separate marking for agent
vs.\ experiencer ``subject.''  Egophoric marking might only occur with
volitional predicates (``cook,'' but not ``be sick'', Tibetan).
Direct egophoric marking might be used in a clause where the speaker
has some relevant and intentional involvement (``he's at the
airport-\I{ego},'' e.g., because I took him).  Egophore marking likely
to be more distant from verb stem than \I{tam} or person.

Egophoric marking centrally codes epistemic source or authority (in a
statement about me, \I{ego} is the authority; in a question about you,
\I{nonego} is).  It might also code for control and agency, such that
\I{nonego} marking with first person can indicate unintential acts
(``I ate a bug-\I{nonego},'' i.e., by accident).


\subsection{Valency}
Possible that applicatives can only be used for animate arguments.
Less often, may only be used with intransitive verbs.

It may be that applicatives in general are less frequent in languages
with rich case systems (less necessary).

Antipassive marker may code for humanness (Rgyalrong).  Antipassive
regularly restricted to particular kinds of verbs: ingestion,
interaction, communication, or traditional activities, and may be
incompatible with action verbs.

In languages that restrict causatives in transitives to verbs of
congnition and perception (know > inform), verbs of ingestion and
consumption (eat > feed) may pattern with them.
% https://eprints.soas.ac.uk/5972/1/JAGGAR_BUBA_HAUSA_EAT_DRINK.pdf

% Mine this paper for additional data. http://linguistics.buffalo.edu/people/faculty/dryer/dryer/KeenanDryerPassive.pdf
% 2020 location: https://linguistics.ucla.edu/people/keenan/Papers/KeenanDryerPassiveProofs.pdf
\subsubsection{Passive} Most often synthetic, with the marking usually,
but not always, closer to the verb stem than \I{tam} marking.

If auxiliary, most common intransitives are \textit{be, become,} and
verbs of motion; most common transitives are \textit{get, receive,
suffer, touch;} even \textit{eat} is attested.

Accessibility in promotion of passive: \I{direct object > indirect
object > adjunct}.

Actor most often represented by an oblique: instrumental, locative
(preposition, \textit{by}, ὑπο, etc.), genitive.  Rarely, special
marking is employed.


\subsection{Semantic Types and Their Frames}

\begin{center}
\begin{tabular}{lllll}
\I{affect} (\E{hit, cut}) & Agent & Target & Instrument \\
\I{giving} (\E{give, lend}) & Donor & Gift & Recipient \\
\I{speaking} (\E{speak, tell}) & Speaker & Addressee & Message & Medium \\
\I{thinking} (\E{consider}) & Cogitator & Thought \\
\I{attention} (\E{see, hear}) & Perceiver & Impression \\
\I{liking} (\E{like, love, hate}) & Experiencer & Stimulus
\end{tabular}
\end{center}

In some languages, some \E{affect} verbs may require an inanimate
agent, usually specific, such as food making a person sick,
\textit{etc}.

In very few languages are \I{liking} verbs like \I{annoying} verbs,
where the experiencer and stimulus are in \I{o} and \I{a} roles.
However, extended intransitives, with an oblique experiencer, are a
bit more common.

\I{attention} and \I{liking} may have extended intransitive frames.
Or dative subjects.  Or oblique objects.

\I{want} verbs may have extended intransitive frames.

The types with more than two arguments may have lexical or
construction splits to determine which of the arguments is in the
primary \I{o} slot (``tell'' vs. ``speak (French)'' vs. ``say'').

For \I{give} there may be i) single verb with different constructions,
ii) gift role is always in \I{o} function, iii) recipient role always
in \I{o} function (rare; amusingly, with \I{dat} for the gift role).
Or, ii and iii may have different lexical items.


\subsection{Associated Motion}
In addition to the basic trans- and cislocative, which link a motion
to the main event of the verb, they may code relative time of motion
(prior: \I{do.arriving}, ``when, after going, X''; simultaneous:
\I{do.going}, ``while going, X''; prior: \I{do.and.leave}, ``before
going, X''). They may distinguish simple deictics (``come in order
to'') vs.\ associated motion (``come while doing'').  There may be a
distinction between ``go'' and ``come'' vs.\ ``go back'' and ``come
back.''  Arrernte distinguishes a ``quickly'' or hurried series, as
well as ``do coming through.''  No deictic center may be present for a
small subset (e.g., ``do here and there'').  In some systems, the
elaboration for \I{s/a} is reduced, but there is a series for \I{o} on
transitive verbs,\footnote{With deixis oriented to the verb subject.
Not sure if that's universal.} ``I saw-\I{go(o)} a man'' for ``I saw
a man going away from me.''  Finally, the target of motion may be
coded for stability (temporary or permanent,
``enter.\I{come.temporary}'' ``come in,'' vs.\ ``sit.\I{go.perm}''
``go and sit and stay there'').  

Very rarely the two core associated motion affixes may be used
together for ``away and back'' sense.

Expansion of the basic system often includes ``up'' and ``down''
senses (and these may be the only extensions); and then ``in'' and
``out.''  Erromangan has a misdirective, ``away from expected
direction, off to the side,'' etc., and Northern Paiute includes
random motion in the set.

Rarely these affixes may be confined to imperatives.  Affixes might be
different depending on transitivity of verb, number of subject, or,
very rarely, tense or aspect (probably from \I{svc}s).
% Data Handout for A Cross-Linguistic Survey of Associated Motion and
% Directionals, Daniel Ross

These locate an event in space much like tense locates an event in
time.  Though using these may add locative arguments to clause, the
main verb is always foreground.  Further, they may link discourse,
with things like ``he came, and he argued-\I{cis} with me,'' with
``come'' and \I{cis} marking the same path information.

Directional affixes may be grammaticalized in various ways (inward =
perfective, downward = progressive, upward = imperative; cis =
inceptive, change of state, trans = endpoint of activity); or more
metaphorical senses (down = not fully satisfactory, up = better, back
= return to health or satisfactory state).

Like tense and aspect, these locatives may be restricted or forbidden
in subordinate or nominalized clauses (these are inflectional, not
derivational). 
% https://ecommons.cornell.edu/bitstream/handle/1813/13037/Deal.pdf?sequence=2

Associated motion appears to occur in about 1/3rd of languages, with
certain hotspots (Australia, the Andes and Amazonia).  It is less
likely in languages with serial verbs.  The forms appear to be
diachronically unstable, with potentially wide variation between
otherwise closely related languages.

% http://linguistics.berkeley.edu/~fforum/handouts/vuillermet_AM_fforum_handoutFINAL.pdf
% http://www.academia.edu/7866301/Reconstructing_the_category_of_associated_motion_in_Tacanan_languages_Amazonian_Bolivia_and_Peru_
% https://escholarship.org/uc/item/85r367r3#page-40
% http://www.ddl.ish-lyon.cnrs.fr/trajectoire/23us23efd5ps/Presentations/Pres_Guillaume091107.pdf

\subsection{Participles}
% Ksenia Shagal, ``Towards a typology of Participles,''
% https://helda.helsinki.fi/bitstream/handle/10138/177418/Towardsa.pdf
Participles may have an \textit{inherent} orientation or a
\textit{contextual} orientation.  Most European languages have
inherent orientation, where the role participant is determined by the
participle form (active participle for agent, passive for patient).
In contextually oriented participles, several different participants
(agent, patient, location) can all be encoded, and usage determines
the interpretation.  The relative accessibility hierarchy will be in
play here, too.  Contextual oriented participles are generally
generous in what they will accept on the hierarchy.

\rara{Separate inherent orientation participles distinguishing \I{s}
from \I{a} role.}

Locative orientated participles (``live.in-\I{pcpl} city'') and
instrumental orientied participles (``book write-\I{pcpl} pen'') are
possible, but rare-ish.  Instrumental may be interpreted as reason,
attaching to a generic noun meaning ``way'' or ``time.''

It is possible to have passive participles (i.e, oriented on \I{p}) in
languages which do not have finite passive constructions at all.  It
might be obligatory to state the agent in such constructions.  Passive
participle forms are likely to be the form used to relativize further
down the accessibility hierarchy.

\I{s+p} absolutive participles may only be permitted for low-agency
intransitives (``fallen leaf'' but not ``danced woman'').  Telicity
sometimes a core consideration.  However, participle patterns may be
have verb-specific restrictions.  In Finnish, ``like'' can relativize
the stimulus (``the music one likes'') but ``hate'' and ``be afraid
of'' cannot.
% "Lexical Meaning Of A Verb As A Factor Shaping Relativizing
% Capacity:"  https://tuhat.helsinki.fi/ws/portalfiles/portal/161945352/shagal_negative_ptcp_30032021.pdf

Participle forms often have several functions in addition to
adjective-like attribution: adverbial (``watching the children,
...''), clausal argument (``I saw he was not running''), action
nominals (``running is good'').  Participles and nominalizations may
be hard to distinguish formally.

Sometimes participle markers include \I{tam} information, sometimes
not.  Markers that do not include \I{tam} information are sometimes
restricted in what verbal \I{tam} marking they may occur with.  Or,
separate \I{tam} markers are used with particples.  When the
participles code \I{tam}, it is normal for fewer distinctions to be
made.  Finally, participles may make no \I{tam} contrasts at all.

There may be separate negative participle forms.  Also, particular
participle forms may not be permitted with negation at all, requiring
a clause to be recast into a particular participle form to be negated
(such as the absolutive participle in Georgian).  Nominal negating
constructions may be preferred.  Rarely, it might not be permitted to
negate any participle types.

Arguments to participles may be as for verbs, or as for verb
nominalizations (genitives frequent).

Even languages without an adjective class can have things quite like
participles, which are used in relative and attributive constructions. 

Participles in some languages are the only way to manage a relative
clause.  Or, some marker plus the participle forms a separate relative
construction. 


\subsection{Converbs}
Moderately more likely in SOV languages. Most things called converbs
require same subject, but an argument can be made for inflected
converb-like functions being part of the same phemomenon (cf.\ Coptic,
Wolaitta).  Forms not inflected for person may or may not require same
subject, though same subject is most common.  Argument marking may be
identical to a main clause, or different.  There may be restrictions
on negation.  Rarely, converb clause may occur at the other margin of
the clause, often with iconic restrictions, as in a purpose or
intention converb allowed to follow the clause, while imperfective may
not (in an SOV language).

Simplest functions: aspectual, imperfective (simultaneous) and
perfective (anterior, usually). Some languages have many forms (all
from Northern Akhvakh): locative (may take location case marking),
inceptive (``from the moment X-ing began''), immediate (``as soon
as''), anterior (``before X-ing''), imminent (``just before X-ing''),
non-posterior (so fast that the converb event may not have time to
occur at all, ``come down here-\I{cvb} before something bad
happens''), conditional, concessive (``although''), similative (``in
the same way as''), gradual (``the more..., the more...,'' with the
main clause just normally marked), cause (``because''), purposive
(``in order to'').

In languages with rich converb inventories, some may be highly
restricted with respect to the main clause verb: Akhvakh progressive
converb only occurs with ``be, remain, see, find.''

In Turkic languages (and some others), converb forms are the \I{lex}
in auxiliary constructions.

% http://www.deniscreissels.fr/public/Creissels-adv.sub.Akhv.pdf
% http://pj.ninjal.ac.jp/vvsympo/NINJAL2013_Shluinsky_handout_changes.pdf
% http://www.livingtongues.org/docs/AVCsOT1.pdf

\section{The Adverb}
In languages otherwise without a well-defined adverb class, or a
regular way of forming adverbs, there may be root adverbs for the
domains \I{speed} (``quickly, slowly'') and \I{value} (``well,
poorly, bad''). In addition to root adverbs, these can show up as
derivations from adjectives which may differ from other adverb-like
constructions somehow (such as being zero-marked). These two semantic
cores are rarely the only one in any particular language.
%http://www.ling.su.se/polopoly_fs/1.99116.1346331417!/menu/standard/file/Hallonsten_Halling_Pernilla.pdf


\section{Expressives}
In some languages ideophones are highly marked, and stand outside
normal phrase structure.  In others, they are more integrated into the
rest of the grammar.

Van Hoey's hierarchy: \I{i} sound < \I{ii} motion / vision < \I{iii}
somatosensory < \I{iv} inner senses.  Within \I{i} are a distinction
between \I{human}, \I{animate}, and \I{inanimate} sounds.  The
starting point for the \I{ii} visual field is motion in general, with
further details on \I{configuration, position, size, shape,
brightness, color}.  \I{shape} is the core of that, with the the
graph expanding out in either direction from there.  The \I{iii} group
is less patterned, containing \I{pain, temperature, kinesthesia,
touch, smell, taste,} with \I{touch} (texture) the core.  Group
\I{iv} is also less patterened, with \I{inner feelings, evaluation,
time}.

Ideophonic synesthesia is not uncommon, with groups \I{i} and \I{ii}
in particular taking on senses under \I{iii} and \I{iv}.  The sound
and motion ideophones seem to be the richest source.
%  Van Hoey, Thomas. (to appear). A semantic map for ideophones. In
%  Thomas Fuyin Li (ed.), Handbook of Cognitive Semantics. Leiden:
%  Brill. 
% Preprint doi: 10.31219/osf.io/muhpd

A cross-linguistic comparison shows modest tendencies for certain
phoneme features to pair with ideophone domains: [+telic] often /k ŋ j
x c g/ or [+labial] and any vowel type; [+wind] and [+friction] often
have [+airflow]; [+motion] has [+labial] or [+tongue resting] (i.e.,
non-coronal), and often [+unround] vowel; [+appearance] has velar.
% Articulatory features of phonemes pattern to iconic meanings:
% evidence from cross-linguistic ideophones 2021.

\section{Number}
The distinction between ``paucal'' and ``plural'' is context
dependent. 

If there is a trial, its use may signify salience of the number, with
the plural being used for three much of the time.

Number suppletion in verbs: not common, aligns ergatively.  Usually
only 1--4 verbs take it, but may reach a couple dozen.  If there is a
sg/du/pl distinction, at least some verbs will just have sg/pl (or
non-pl/pl).  Among intransitives, the posture verbs are most commonly
suppletive, \textit{sit, stand, lie,} and most likely to show numerous
number distinctions (du).  Next: \textit{enter, go, be big, die/be
  dead, hang, arrive, run, come, fall, cry, be little}.  Transitives:
\textit{kill, put/place, throw, give, break, take, bring, carry}.


\subsection{Countability Classes}  In languages with a
singulative-collective distinction the hierarchy is: substances <
aggregates < individuals; with individual most likely to be singular
by default, with a marked plural.  In languages with singulatives,
aggregates most likely to take those.

Singulatives are rather common with small fruits and vegetables, small
animals, sometimes paired body parts, and groups of people (Welsh:
moch \textit{pigs} vs.\ mochyn \textit{pig;} mwyar
\textit{blackberries} vs.\ mwyaren \textit{blackberry}).
%https://dlc.hypotheses.org/1554


% \section{Exponence}
% https://www.academia.edu/41951336/A_Typological_Approach_to_the_Morphome_-_PhD_dissertation

\section{Constructions}
Any schematic construction may, like lexical constructions, be
polysemous.  For example, the English ditransitive:

\begin{grammarlist}
  \item \I{X causes Y to receive Z} (central sense), ``Joe gave Sally
    the ball.''
  \item Conditions of satisfaction imply \I{X causes Y to receive Z},
    ``Joe promised Bob a car.''
  \item \I{X causes Y not to receive Z}, ``Joe refused Bob a cookie.''
  \item \I{X acts to cause Y to receive Z} at some future time, ``Joe
    bequeathed Bob a fortune.''
  \item \I{X enables Y to receive Z}, ``Joe permitted Chris an
    apple.''
  \item \I{X intends to cause Y to receive Z}, ``Joe baked Bob a cake.''
\end{grammarlist}

Similarly, a particular language may have one construction for a
particular job, such as a ditransitive construction, but others have
several with different pragmatic or pivot significance (``I gave him
the book'' \textit{vs.} ``I gave the book to him'').


\subsection{Grammatical Relations \& Alignment}
% Bickel, ``Grammatical Relations Typology''
Grammatical relations are sets of arguments that are treated the same
way by some contruction, such as being assigned the same case or
causing the same kind of agreement.

Nouns and verbs may have separate alignments, such that nouns may
align ergatively but the verbs nominatively, for example.  Grammatical
relations are construction-specific.

\I{s} = intransitive subject, \I{a} = (di)transitive agent, \I{o} =
transitive object (often this is \I{p}), \I{t} = ditransitive theme,
\I{g} = ditransitive goal (or ground).

Rarely, transitive \I{a} is distinguished from ditransitive \I{a} (and
not by a separate case, but by a separate construction).

\begin{grammarlist}
  \item \{\I{S}\} intransitive subject, nominative
  \item \{\I{S, A}\} subject, nominative; accusative alignment
  \item \{\I{A}\} transitive subject, ergative
  \item \{\I{O, T}\} direct object, accusative; indirective alignment
  \item \{\I{O, G}\} primary object, dative; secundative alignment
  \item \{\I{T}\} secondary object
  \item \{\I{G}\} indirect object, dative
  \item \{\I{S, O, T}\} absolutive; nominative; ergative alignment
  \item \{\I{S, O, G}\} absolutive; nominative; ergative alignment
\end{grammarlist}

\I{o} arguments may map to different relations based on: animacy,
humanness, definiteness, specificity or saliency.  In Nepali, animate
\I{o} are marked as \I{g}.  In Swahili, \I{o} is marked on verb only
if animate and/or known.  Or, non-salient \I{o} may be incorporated
into verb.

\I{a} arguments are the inverse of \I{o}: the lower the salience or
animacy the more likely they are to get overt marking, so that a human
may get nominative, a stone ergative.  In some languages inanimate
\{\I{S, A}\} is basically impossible, with voice changing or
incorporation to handle things like ``the wind broke it.''

\begin{grammarlist}
  \item speech act participant > kin/name > human > animate >
    inanimate > mass
  \item specific > nonspecific referential > generic/nonreferential
  \item known/topical/thematic/definite > new/focal/rhematic/indefinite
  \item singular > plural
\end{grammarlist}

Role marking may be sensitive to other roles within the clause.  Yurok
only marks \I{o} when there is a \I{3p} subject.  Sahaptin only marks
\I{A} if the \I{o} is a SAP, Tauya only if \I{o} is human.

It is unusual for languages to have obligatorily filled grammatical
role positions (very much unlike English).

In Eastern Kiranti, a specific \I{o, g} argument can be an \I{np}
(adjective attribution, etc.), while nonspecific must be a bare
\I{n}.

Floating quantifiers may prefer a particular grammatical relation.  In
Tagalog, \textit{lahat} ``all,'' floats to Wackernagel, but goes with
the proximative relation.  In Yélî Dnye, quantifiers float to the
preverbal position, but go with \{\I{s, o}\}

Focus constructions may be different for different relations.

The non-ergative clauses in so-called split ergative systems are
rarely nom-acc (\I{a = nom, p = acc, s = nom}).  Instead one can get
\I{a = abs, p = abs, s = abs} (common), \I{a = abs, p = obl, s = abs}
(common) or even \I{a = erg, p = abs, s = erg} (common in Mayan).  The
aspect splits often reflect (historically) intransitive \I{aux}
constructions.  Thus, in rare circumstances, one may get splits in
future clauses, or with negation, if they reflected an original
\I{aux} construction.


\subsection{Complement Clauses}
These come in three types.

1.  Fact type, indicate that something did take place.  Similarly
marked to main clause.  If subject is the same across clauses, it
isn't likely to be omitted.  Usually marked with some complementizer
element which will have other functions in the language (often ``say''
or ``be like'').  Complementizer may code reliability (sure fact
vs. possible fact).

2. Activity type, indicating extension in time.  Often similar to a
noun phrase, but will still have a subject.  If the subject is the
same, it may be omitted.  Or the verb may have a special form.
Generally less specified in TAM and negation than a main clause; may
not include same bound pronominal elements.

3. Potential type, typically less like a main clause than the Fact
type and less similarity to a NP than the Activity type.  In some
languages, the subjects must be the same.  Reduced TAM and pronominal
marking.  Implicit type reference to same or posterior time.
Generally a special verb form (``infinitive'') or may take marking
similar to dative or some other case.

Languages may range from one to 5--7 complement types, with subtypes.

Attention verbs (``see, hear, show'') typically take Activity
complement.  May take Fact, for completed actions or of state.
``Find, discover'' are expected to take Fact.

Thinking verbs (``think (of, about, over), consider, imagine'') take
Fact, or sometimes Activity (``think about'').  ``Assume, suppose''
take Fact.  ``Remember, forget'' take Fact, with English unusual in
taking potential (``I remembered to shut off the stove'').  ``Know,
understand'' take Fact or Potential.

Deciding (``decide, resolve, choose'') take Fact or Potential.

Liking (``love, prefer, regret, fear'') take Activity or sometimes
Fact.  ``Enjoy'' takes Activity.

Speaking has several subtypes.  ``Say, inform, tell'' usually take
Fact.  ``Report'' takes Fact or Activity.  ``Describe, refer to''
takes Activity.  ``Promise, threaten'' takes Potential, which may be
in the indirect object slot.  ``Order, persuade'' generally take
Potential. 

\subsubsection{Epistemic Stance}
In addition to things like the subjunctive, the epistemic stance
towards an embedded clause can be affixes, clitics, particles,
adverbs, and auxiliaries.  Jacaltec has different two complementation
conjunctions for verbs of speaking, one indicating certainty, one
indicating reservation or hearsay.
% https://lsa2009.berkeley.edu/alt8/vanlier_boye_abstract.pdf

\subsubsection{Expletive Negation}
% ``Fear complement clauses in typological perspective,'' Dobrushina,
% Nina, 2017.
In fear clauses, especially if not indicative, negation may be a
requirement of the construction, though it is not negating the fear
clause.\footnote{French has it optionally: \E{je crains qu'il (ne)
    vienne} ``I'm afraid he might come.'' vs.\ \E{je crains qu'il ne
    vienne pas} ``I'm afraid he won't come.''}  If a language allows
different complement constructions for fear-verbs, one may take
expletive negation while the other does not (cf.\ Russian \E{čto}
vs.\ \E{kak...by ne}).

These may evolve from wish constructions, or from negative
purpose clauses.

When fear clauses look like embedded questions, even if indicative,
expletive negation is more likely.

Fear-verbs likely to take this: \E{fear, prohibit, hinder, prevent,
  avoid, deny, refuse}.  In a few languages positive predicates
(\E{hope}) might have some constructions with expletive negtation.

Expletive negation may also occur in: exclamatives, emphatic
questions, concessive conditionals, before clauses, until clauses,
polite requests, comparatives.


\subsection{Desententialization}
The most desententialized is purpose clauses: purpose $<$ before $<$
after, when $<$ reason, reality condition.  Also: phrasal, modal $<$
desiderative, manipulation\footnote{\textit{urge, suggest,} etc.} $<$
perception $<$ propositional attitude,\footnote{\textit{believe} and
the like} knowledge $<$ utterance.

% Add more from: http://www.academia.edu/2099252/Argument_coding_and_clause_linkage_in_Australian_Aboriginal_languages

\subsection{Temporal Clauses}
% Temporal Adverbial Clauses in the Languages of the World: Clause-
% Linking Strategiesm, Jesús Olguín Martínez.
In addition to conjunctions (which might be bound, or clitics),
temporal clause may be signaled with temporal nouns (``the day I
left''), as well as relativized temporal nouns (Mandarin), verbs
meaning ``finish'' (``I woke up, it was finished, I left'' = ``and
then''), quantifiers (``all my leaving, my feeling sick'' = ``as soon
as I left''), serial verbs (``go reach'' = ``until''), etc.  Temporal
nouns might be case-marked with some locative case (but usually not),
or rarely doubled.  Non-generic temporal nouns (``day, year,'' etc.)
often used.

Either figure (main) or ground (subordinate temporal) might get
marking for a sentence type (``after he got sick, he died'' vs. ``he
got sick, he died-\I{seq}'' or the like); or both may be marked.

\I{when}-clauses are often polysemous with conditionals, especially
for generic or habitual senses (``when kept in the heat, flowers
wither'').  Sometimes also polysemous with \I{while, because, after}.
\I{when}-clauses are not specific about the time (both sequence and
duration), relying on context.  Rarely, the ground (when) clause may
take some overt device, and the figure must also take some marking,
usually ``and.''  Can be nominalized clause, with or without an
article. 

\I{while}-clauses may be marked with explicit marker, or \I{ipfv} (or
related) aspect.  Temporal nouns may be used (``time'', or, in Western
Europe ``length'').  They often develop concessive meaning, as well as
conditional, cause, and \I{after}.  Asyndeton not uncommon device.
Deranking is also common (converbs, etc.).  Locative or comitive case
markers can be used.  Sometimes used as complement clause of verbs of
perception (``I saw while the dog was running'').  Generic temporal
noun less common, though ``duration, amount'' are seen.  Correlative,
with identical conj.\ in each clause also seen (also: ``first, still,
also'').  

\I{after}-clauses may be marked with adverbial subordinator (related
to ``first'' possible), sequencing verb forms (converbs, say), and
``and then'' devices (which marks figure).  May also signal \I{when,
while, if, because,} though \I{after} is much less often polysemous.
If asyndeton, in iconic order.  Deranking frequent (converb- or
consecutive-like), often marking figure not ground.  Ablative usual
for nouns or nominalizations.  Nouns unusual, usually ``back.''

\I{before}-clauses tend to have overt subordination device.  Even in
languages with an overt device, negation is often involved (sometimes
optionally, sometimes not; sometimes codes ``before'' vs.\ ``until'').
If \I{before} is lacking, ``not yet'' may be used (``he not yet die,
he said this,'' = ``before he died, ...'').  Deranking common, and is
likely to be polysemous.  May also signal \I{while, in order to}
cause, though polysemy here less common with overt adverbial
subordinator.  May be encoded by ``not yet.''  Ground clause may not
happen at all, and a few languages encode distinction, or the
differences is did happen vs.\ uncertainty.

\I{until}-clauses, marking terminal boundary, tend to be marked with
adverbal subordinators.  Bound subordinator less common.  Deranking
fairly common.  Negation possible, though less so than with
\I{before}.  ``While... not'' is a regular circumlocution.  Asyndeton
possible but rare, with iconic clause order.  Reduplication of verb
(``man gnaw gnaw kava break''), but a verb meaning ``go'' might be
reduplicated.  Derived from ``arrive, go, reach, touch.''  Serial
verb, ``go reach'' in Oceanic languages, but also in a few others.
For nouns, ``end, limit, edge, border'' are seen.  ``Only, just''
might be used.  Correlatives seen but unusual.  May be related to
\I{as a result, while, before, if}.

\I{as soon as} clauses, usually have unbound adverbial subordinators,
or deranking.  Adverb ``immediately'' also a device.  In West African
languages, verb doubling is common in \I{as soon as} (``arrive Koku
arrive Asiba leave'').  Outside Africa, the verb doubling may be a
\I{v v-neg} construction (Japanese, ``arrive \I{q} arrive-not
\I{q}...'').  Correlative (``no sooner... than'') not uncommon.

\I{and then} constructions might encode time between situations (Khwe
has three: short time, 1--2 days, more than 2 days). Immediate is
possible.  Rare: might be two, one which is the same scene, on which
changes scene.


\subsection{Negation}
% http://youngconfspb.com/application/files/2115/1334/8210/Miestamo.pdf
Negative affix only a bit less common than negative particle.  Very
rarely, might be circumfix.

Asymmetric negation (different structure under negation beyond mere
presence of marker) reasonably common, with a mix of symmetric and
asymmetric within a single language a bit more common than either
alone.  Possible asymmetries: lexical verb loses finiteness (sometimes
with \I{aux} negator, sometimes other \I{aux} with nominalized,
negated \I{lex}); negatives use irrealis of some sort; negatives have
some sort marking used for asseveration in non-negated; difference in
grammatical categories (TAM, argument structure; Harar Oromo removes
all person marking distinctions in negative).  Negation is stative, so
negated clauses may partake of some stative construction behavior.
The discourse context of negatives \E{contra} affirmatives necessarily
means the details matter less.

Nonstandard negative constructions more likely in: imperatives,
existentials, nonverbal clauses.

Negative imperatives might be (with WALS counts): usual imperative +
usual negation (23\%); usual imperative + different negation (37\%);
different imperative + usual negation (11\%); different imperative +
different negation (29\%).

Circumfix all over planet, just not common.

Possible distinctions, beyond simple ``not:'' noun vs.\ non-noun, verb
vs.\ non-verb (where adj.\ may pattern with either); different negator
for the future; prohibitive; participle negation; negative particle
for non-existence (which may be like ``no'' or plain negation for
nouns).  Non-verb negator can be grammaticallized from ``other,
different; refuse, not want.''

\subsection{Questions}
Focus questions (``do \textit{you} want it?'') usually look like polar
questions, but might have a special affix or structure; could have
affinity with alternative question construction.

Rarely, negative polar questions might be distinct from positive polar
questions. 

Having unmarked polar questions is extremely rare, whether lacking an
affix declaratives do have (Sheko) or being identical in form and
intonation to a statement (Yeli Dnye).  Most common is a marked polar
question and unmarked declarative, though marked declaratives can
happen (Crow, Sabanê).

Intonation is among most important ways to mark questions, though it
usually is accompanied by other markers.  Less than 20\% of one survey
(of 955 languages) use intonation alone.  The rising pattern, as in
English, is in no way universal.  In West Central Africa, an areal
feature of questions involves falling pitch, vowel lengthening, and a
breathy end to the intonation unit.

Word order change to mark questions is extremely rare, basically
confined to Western Europe.

\rara{Koasati uses a glottal stop infix as question marker (usually
morpheme boundary).}

\rara{Halkomelem uses auxiliary verb for polar questions.}

\subsubsection{Answers}
Some languages have words for ``yes'' and ``no'' for replies to polar
questions, and a separate pair for response to commands or requests.
Repeating clause parts (verb especially) is normal in languages
lacking yes/no.

% A Typology of Questions in Northeast Asia and beyond, Hölzl, 2018.


\subsection{Conditionals}
The protasis of a condition and formal marking of the topic may be
identically or similarly marked (including ``if X \I{adp}'' or the
like for topic marking).

\subsection{Concessive Conditionals}
These come in three types, with several coding types available.
Frequently polysemous among themselves, and with plain conditional.
Conditional > concessive conditional > concessive.  Less polysemy in
systems with more finite constructions, much more polysemy with less
finite constructions.

\textbf{Scalar.} ``Even if it rains, we will go outside.''  Might be
identical with conditional; conditional clause plus focus particle
(``even/also''), like English ``even if,'' Japanese ``V-te mo;''
polysemous subordinator ``even.if/though;'' or special subordinator.

\textbf{Alternative.} ``Whether it rains or not, we will go outside.''
Can be based on balanced conditionals (if... if...); embedded
interrogatives (German ``ob... oder...''); with balanced irrealis;
with ``you want'' (Turkish ister V, ister V); expression of
irrelevance (Mandarin ``no.matter... or...'').

\textbf{Universal.} ``Whatever the weather is like / no matter how much
it rains, we will go outside.''  Generally formed with question word
as part of construction.  Can be particle attached to verb (Quechua
``\I{wh} V-\I{cond}-even'') particle following \I{wh}, such as
\I{wh}-even; particle before \I{wh} (Nahuatl ``only \I{wh}''); with
irrealis (Turkish [\I{wh V-cond V-subj}]); expression of irrelevance
(Mandarin ``no.matter \I{wh}'').

% Talk slides: https://www.youngconfspb.com/application/files/8115/7532/5100/Bossuyt-Concessive_conditionals.pdf

\section{Discourse}

\subsection{Definiteness}
In languages with (in)definite articles, where indefiniteness is
marked by articles within the \I{np}, indefinite nouns tend to be
unrestricted in core argument position.

In languages without (in)definite articles, non-specific arguments are
more likely to have restricted core case functions, and definiteness
distinctions tend to be encoded in the verb phrase, generally with
various valency operations (\I{dom}, incorporation, antipassives,
etc.).  In such languages, existential expressions are preferred to
introduce novel discourse topics.  Even a language with articles, such
as French, may disprefer indefinite subjects, using an existential +
rel.\ expression in colloquial speech.  If semi- or de-transitivizing
verbs forms are available in an article language, they might be used.

Subjecthood cline: definite > generic > specific indefinite >
indefinite.  In some languages, subjects \textit{must} be definite
with event verbs.

Objecthood cline: definite > specific indefinite > generic >
indefinite.  Again, indefinite objects may trigger valency reducing
forms, making objects oblique (or incorporated).

Individuation may be part of the clines above, with, for example,
``the stars'' as a mass treated more like an indefinite than ``the
moon.'' 

% https://hal.archives-ouvertes.fr/hal-02881062/document

\subsection{Topicality}
Generic topicality hierarchies: speaker > hearer > \I{3rd}; human >
animate > inanimate; agent > dative > patient; large > small (and
adult > child); possessor > possessed; definite > indefinite; pronoun
> full \I{np}.

Referential hierarchies: Speech-act participants > Kinship/Name >
Human > Animate > Inanimate; Specific > Non-specific referential >
Generic; Known/Topical/Thematic/Definite > New.  Different languages
take different approaches for SAPs, with both 1 > 2 and 2 > 1 found
(the latter a politeness matter, apparently) and number may play a
role, too. Found in the wild: \I{1pl/2pl > 1sg > 2sg}, \I{1pl > 2},
\I{2pl > 1 > 2sg}.
% http://elanguage.net/journals/lsameeting/article/viewFile/2826/pdf

\subsection{Focus}
Focus constructions may differ by argument type, clause type,
polarity, etc.

Focus markedness (from least to most): indefinite \I{np}, definite
\I{np}, pronoun, bound pronoun, zero.

\I{wh}-questions generally require a focused answer.  Further, if the
language has \textit{ex situ} \I{wh} words (initial, pre-verbal, etc.),
the focus phrase will usually end up in the same position.

Clauses with focused elements tend not to have overt topics.

Focused subjects may move into object position. In an \I{sov}
language, the focused subject may move immediately before the verb.
In a very few languages, subjects may be incorporated into an
intransitive verb to focus them.
% https://spot.colorado.edu/~michaeli/courses/LAM7420/Lambrecht.pdf

In some (\I{sov}) Omotic languages, what look like the subject
agreement affixes (clitics?) can move around and attach to a focused
element.  With content questions, the often fronted question word
always gets the subject.  In Huallaga Quechua, the evidential marker
goes on the focus (or the verb, in the most neutral, predicate focus
situation).

\subsection{Functions} Discourse particles, expressions, and adverbs
serve a set of core discourse functions.
% http://www.ling.uni-potsdam.de/~stede/Papers/StedeSchmitz00.pdf

\textbf{Structure.} What are usually called \I{push} and \I{pop}
introduce and recover from minor digressions, ``by the way, but,
anyway.''  The \I{check} function asks the listener to verify, ``isn't
it?''  There are two editing functions, \I{repair} which fixes
information, ``no, rather...,'' and \I{exemplify} which gives an
example, ``for example, for instance.''  The \I{hesitate} function is
filler to let the speaker figure out what they're going to say next.
\I{uptake} indicates that the speaker has heard and understood the
listener, and is ready to move on, ``so.''

\textbf{Coherence.} These functions address knowledge assumptions and
attitudes, with \I{known} indicating assumed shared knowledge, ``as
you know, indeed,'' and \I{revised} indicating an updated assumption,
``after all.''  A speaker might restate something ``in other words.''

\textbf{Attitude.} \I{positive} and \I{negative} ``unfortunately.''
\I{indifferent} ``as far as I'm concerned.''  \I{surprise} ``did you
actually?'' and \I{prefer} ``I would rather.''  The \I{smooth}
function can generate a vast stream of words which tone down, lighten,
and moderate face threats.  And \I{emphasize} marks something as high
on a scale, ``really.''

Discourse particles in practice might indicate only one of the major
functions, or might, depending on use, cover several.  Alternately,
several functions might be joined into a single discourse construction
that expresses all those functions simultaneously.  A discourse
construction may have a particular function interpretation only
questions.

Different function constructions may be possible in different places
in a discourse: initiation of a turn, reply (or reply-initiate) to a
turn, midclause, or uttered alone, with different nuances in meaning
possible. 
% https://www.researchgate.net/profile/Salvador_Borderia/publication/348190527_Using_discourse_segmentation_to_account_for_the_polyfunctionality_of_discourse_markers_The_case_of_well/links/5ff32d1b92851c13feeb11bc/Using-discourse-segmentation-to-account-for-the-polyfunctionality-of-discourse-markers-The-case-of-well.pdf

Discourse construction frequency and selection may vary by register,
genre, politeness, etc. 

\subsection{Allocutivity and Exaltation} Encoding some salient feature
of the addressee---\textit{other} than as a predicate argument--- in
the clause can be called allocutivity.  Can be: gender, number,
attitude (respectful, familiar), solidarity.  In Japanese, the
\textit{-masu} family which encodes respect.  Basque (across different
dialects) encodes the gender and sometimes number of the addressee.
In Basque, the allocutive endings are not permitted when the addressee
is also a core argument of the verb (they seem to have evolved from
`ethical' datives), but other languages do not seem to have this
restriction.

In Pumé the allocutive encoding often blends with person marking.  A
language that does not encode gender of arguments might encode gender
of addressee.

The English interjection ``yo'' can be considered a kind of allocutive
encoding (it suggests familiarity).

Allocutivity is different from subject- and object-exalting honorific
verb forms, such as are found in Japanese, or the rich multi-class
honorific vocabulary of Javanese and the referent-exalting vocabulary
of Tibetan.

Allocutive marking generally confined to independent clauses, but
might be further confined to positive, declarative clauses.  Javanese
allows allocutive marking in relative and some complement clauses, but
this is seen as rather exaggerated politeness, and appears to be
confined to clauses in which the addressee was also a participant.
Imperatives, sometimes yes, sometimes no.

% https://www.academia.edu/1817520/Verbal_allocutivity_in_a_crosslinguistic_perspective


\section{Lexicon}
Consider a few phonesthemes.

Diminutive/medial/augmentative may code gender of the speaker (Weining
Ahmao, in the classifier system).

Special formal/elevated vocabulary word shapes may have several
patterns of systematic relationship between it and base level words
(Javanese).

% http://www.linguistics.ucla.edu/faciliti/wpl/issues/wpl17/papers/40_polinsky.pdf
\I{vso} languages are more likely to have even noun-to-verb ratios
(lots of verbs derived from nouns), while \I{sov} languages are more
likely to have \I{n+v} idioms taking up the slack, resulting in more
nouns.  This includes \I{svo} languages that lean \I{ov}.  Regardless,
there are almost always more nouns than verbs (though perhaps not by
many).

Respect forms may contain morph for ``lord'' or ``sky.''

Some Austronesian languages have an ``anger'' vocabulary, not
swearing, for verbs and nouns.  Some common deformation patters.
% The Angry Register of the Bikol Languages of the Philippines:
%  https://www.sil.org/system/files/reapdata/18/02/03/18020356736639094334766403643144945625/Lobel.pdf

Hunting, fishing or territorial languages: to conceal your intent from
animals or spirits, or with tabu considerations; circumlocution and
non-systematic deformation common practices for this.

May be separate interjections to: summon an animal, disperse an
animal, urge to start or stop work (and forward, backward, slow down,
move right or left).  Less common, to stop (or start) chasing or
fighting, or an urge to eat/drink.  Calming sounds or interjections
are also possible.  Summon, disperse, and work calls may be distinct
by animal (not a huge number of distinctions, though).  Some calls may
be restricted to herds or flocks.  Most variety in calls for domestic
animals, but dispersal possible for wild animals.  Likely to be
phonologically unusual, with things like [ʃːː] possible, as well as
paralinguistic sounds, or clicks.  A sing-song intonation possible.
Summons are slow, with possibly repeated word; dispersals have high
articulatory speed, and sibilants are common.  ``Stop'' commands may
have prolonged vowels or consonants.  Can be up to a few dozen in
total.  Generally isolated but possible to use adverbs of speed, and
expressions of location.
% "Conative animal calls in Babanki and Bum:"
% https://hal.science/hal-04132552v1/preview/Conative%20animal%20calls%20in%20Babanki%20and%20Bum%20.pdf
% "Conative calls to animals: from Arusa Maasai to a Cross-linguistic
% prototype." 

Body parts become special concepts in stages: 1) a region of the human
body, 2) a region of an inanimate object, 3) a region in contact with
an object and 4) a region detached from the object.  The landmark path
(``extremity, peak'' $>$ ``head'') goes in reverse.
% http://www.sciencedirect.com/science/article/pii/S0388000112000423

Over time words move from external and objective to subjective and
grammaticalized (``boor'' from farmer to oaf, ``feel'' from touch to
experience emotion, ``insist'' from perservere to demand to believe
strongly).
% https://www.reddit.com/r/linguistics/comments/34r1vk/where_can_i_find_a_list_of_metaphors_that_have/

Once a conceptual metaphor has taken root (\I{seeing} is
\I{understanding}) vocabulary merely related to the concept of seeing
may also be dragged into the metaphor later.  You don't expect
``brilliant = intelligent'' to happen until the first metaphor is well
established.
% http://www.e-revistes.uji.es/index.php/clr/article/viewFile/1363/1206

\subsection{Movement}
Verbs of rotation may distinguish features: internal vs.\ external
axis; elevation over landmark; control; single vs.\ repeated turn.
% https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2564011


\subsection{The Body}
Many languages have a ``psychic being'' that experiences emotions,
``my psyche became frightened, is happy,'' etc., instead of the entire
person experiencing them.  This psychic being may be an internal
organ: liver, heart, stomach, diaphragm, etc.

Some body parts may be conceived of as parts of other body parts,
where ``my lips'' must be ``the lips of my mouth.''

Some languages with obligatory inalienable possessives on body part
nouns may have a special ``derelationalization'' construction.
Tzutujil has a suffix (while possessors are a prefix).

Inessential body parts (``beard, scar'') may be treated differently
than vital ones.  Predicate possession of a vital body part is likely
to have additional interpretations (``I sit with a belly'' = satisfied
with food; ``he really has ears'' = he has good ears).

Body part properties may be expressed: ``on/to W there is an A P,''
``W has an A P,'' ``W is provided with an A P,'' ``W's P is A.''  A-P
(or P-A) compounds are not uncommon, as both properties and exocentric
nouns. 

Instrumental function of body part terms may be marked in unexpected
ways (``go on/by foot; carry in my hand, on my shoulder'').

A body part may be etymologically related to its most obvious or
common function (such as ``fist'' < ``pound'').

A person experiences and interacts with the world via their body.
Expressions describing that may involve the person or the body part in
many ways across languages.

\begin{grammarlist}
  \item The person may be possessor (``wash his face,'' ``he died by
    the hand of the executioner'').  In some languages the person is
    excluded strongly except as possessor (``he hung his neck'' rather
    than ``he hung himself'').
  \item The person may be indicated with an oblique (often dative),
    (``to me the belly aches, I wash to you the hands, I burnt to me
    the hand'').
  \item In topic-prominent languages, the person may be the topic
    (``elephant-\I{top} nose is long'').
  \item The body part may be a locative (``I have a pain in my ear''),
    ablative (``I have a pain from my ear''), or some other location
    construction.
  \item Or, the person may be the locative (``at the child the head
    aches'').
  \item Body part terms may be assumed to be part of the agent or
    subject of a clause (``I opened the eyes'' = my eyes, ``she
    snapped with the finger'' = her finger).
  \item In a few languages, both person and part may take identical
    role marking (``dog-\I{acc} you-grabbed tail-\I{acc}'').
\end{grammarlist}
% https://christianlehmann.eu/publ/lehmann_body.pdf

\subsubsection{Sickness}
Ways to indicate being ill: intransitive predicate (``be sick,
have-a-cold,'' etc.); motion with disease as subject (``fever came to
me'');  possession (``I have a fever''); commitative (``I am with a
fever''); acquisition with disease agent (``illness grabbed me'');
container schema (either ``I am in malaria'' or ``malaria is in me''
available); acquisition with disease agent cannot usually be
passivized.
%https://cadernos.abralin.org/index.php/cadernos/article/view/322

\subsection{Location and Posture}
Some languages are more location precise, some are more posture
precise.  In either situation, there is enough information for a
listner to figure out the answer to ``where is X?'' by insepction.

Unmarked location construction for stereotypical, canonical situations
have several types: Type 0: no verb at all in \I{blc} (Basic Locative
Construction); Type 1: single locative verb (1a copula, as English; 1b
locative or existential verb, as Japanese); type 2 which has a small,
3--7 set of locative verbs; type 3, which has a large set of
dispositional verbs, 9-100, such as Mayan.  A few languages are mixed,
such as Goemai which uses both type 2 and 3 together in \I{svc}s.

\I{blc} hierarchy: animate ground > figure pierced > ground pierced >
adhesion > core scenes (ring on finger > apple on skewer > arrow in
apple > stamp > cup on table / fruit in bowl).  The core scenes are
most common \I{blc} use, with the full hierarcy quite rare (of course,
English).

Non-canonical location will use more complex constructions.  

Locative copula may make animacy distinctions (Japanese).  Or, it may
distinguish from known and unknown (used in questions, or when the
situation is unclear).

Type 2: \textit{sit, stand, lie, squat, hang, be near, be located as a
mass, be supported in a medium (like water);} also dispersion,
\textit{spread, cover,} and attachment \textit{be fixed, be stuck to,
be bound}.  Which verb is used for a particular item may have
complex decision trees invoking not only orientation, but
prominence, animacy, respect, containment (``the dog sits in the
living room'' even if it is standing, due to the room), contact,
duration of state of affairs, size relation between figure and ground.
Default may need to be memorized like gender, but non-canonical
orientation means a different verb can always be used.  ``Sit'' seems
often the default for uncertain situations.  ``Hang'' may get quite
surprising usage (implies permanence, due to attachment, or customary
behavior, such as terrestrial animals going about their normal
business).  May be separate ``stand'' verb for inanimates.

Type 2 verbs tend to grammaticalize: imperfective, progressive (>
present), or become articles.  The default posture verb for an item
may pervade the language across all constructions in subtle ways: Yêlí
Dnye has not only posture verbs, but internal action (\textit{sit
down, stand up}) verbs of putting (\textit{put standing, put
sitting}), verbs of taking (\textit{take sitting X, take hanging
X}).

Contact vs.\ no-contact a common distinction to make among some
adpositions (for example, ``on'' vs.\ ``over,'' ``on (a wall)''
vs.\ ``in front of'' in English).


\subsection{Perception}
% https://research.jcu.edu.au/lcrc/storeroom/research-projects/evidentiality/folder-2-sashas-publications/perception-cognition-2013
Expressions of cognition frequently resort to verbs of seeing,
hearing, smelling, or feeling.  ``Seeing = knowing'' is not as
universal as European languages make it seem, but is still widespread
across various features (evidentials often make the visual evidential
the default, least marked).

A single verb meaning both ``see'' and ``hear'' is possible, with
instrumental (ear) or object nouns (voice, sound) indicating a
``hear'' sense is intended.

Polysemous senses of perception verb \textit{roots} often
disambiguated by different case structure or complementation
constructions.  Perception verbs occasionally align with cognition
verbs in using different case relations.

Perception verbs might reject derivation (causative, passive, etc.).

Imperatives of several perception verbs (and ``know'') used as
attention getting discourse markers.

``See'' might: warning, ``look out, beware;'' ``discover;'' suggest
special knowledge; ``know'' in perfective or stative forms; ``look
after;'' ``have an opinion, judge.'' Or: ``desire.''  Expressions of
socializing might use sight verbs.  The eye may connote (sexual)
desire or aggression in general (extended eye contact expected in some
cultures, but highly aggressive and threatening in others —
communicants might not even look at each other).  Seeing certain
things is a common tabu, but hearing rarely comes in for such tabus.
Sight might be supernatural.
% https://www.degruyter.com/view/journals/cogl/29/3/article-p371.xml

For sight, may be distinct lexical item depending on thing seen,
breaking along: surprising object, new object, old object, fact,
future, experiential perfect (``I had seen this man''), present (more
than one break possible, though not more than three lexical items seen
in paper data).  Some languages may make one end of this continuum the
most common default visual verb, others the other end.
%http://www.youngconfspb.com/application/files/6914/8424/2059/Waelchli.pdf
% Specific and non-specific perception verbs and lexical typology
% Bernhard Wälchli

``Hear'' might: knowing, understanding, remembering.  Often, ``obey.''
May be colexicalized with sense of smell.  The ear can be the organ of
understanding, memory, but also of emotions, intention, and obligation.

Highly elaborated vocabulary for tastes and smells happen, but are not
common.  In the languages that do have richer aroma terms, food smell
vocabulary is common, possibly including cooking method; the smell of
rotting and feces seems to be common; animals, as well as blood and/or
fat in meat preparation; strongly medicinal or herbal aromas; mold and
mustiness.
% https://www.reddit.com/r/conlangs/comments/huedqu/on_smell_terms/

\begin{center}
  \begin{tabular}{lll}
    Experience & Activity & Phenomenon \\
    \hline
    see & look (at) & look, seem \\
    hear & listen (to) & sound \\
    feel & feel/touch & feel (like) \\
    smell & smell & smell (of/like) \\
    taste & taste & taste (like)
  \end{tabular}
\end{center}

There may be different or identical lexical items of any of the types;
or differing by construction (as English for the bottom
three).  \I{s=o} lability (as in English ``taste'') applies even to
sight verbs in some languages.

Positive and negative judgement of perceptions are likely to split for
sight (``beautiful, ugly'') with just ``good'' and ``bad'' for the
other senses much of the time, with occasionally specialized terms
following no particular pattern.

Viberg's lexicalization and markedness hierarchy: sight > hearing >
touch, taste, smell (i.e., more lexical distinctions for sight than
for smell).

% See also: https://helda.helsinki.fi/bitstream/handle/10138/337473/Raatikainen_Oona_MastersThesis_2021.pdf?sequence=3

If there's a lexical difference between experience and activity,
experience verbs will often be more transitive.  In ergative
languages this holds less often, and might even have intransitive
constructions for both.
% Competition and Variation in Natural Languages: The Case for Case
% edited by Mengistu Amberber, Helen de Hoop, p.102

May be internal, ``proprioceptive'' sense constructions, for things
felt happening to oneself (``feel'' vs.\ ``feel internally'' is
attested). 

Distinction between objective (``look'') and subjective (``seem'') (or
non-mediated vs.\ mediated).  May be a single mediated verb across all
senses.
%https://pdfs.semanticscholar.org/cf1d/359040e7c743ede0e63521e748f2867e931e.pdf

A [-control] root such as ``see'' may become [+control] in the
imperative (``look at!'')


\subsection{Pain}
Crosslinguistically few root words to describe pain.  Rather, idioms
from four domains are commonly used: burning (including smolder,
shine, boil), destruction or deformation (sharp or pointed instrument;
break, tear; press, pull; burst), sounds (usually for sensations in
the ears or head), and motion (circular movement, twist, jump).
Atelic verbs may be nominalized to produce telic/stative
constructions.

The body part experiencing pain may be locative (``in my hand''), a
subject (``my hand hurts''), or an agent (``my hand hurts me'').

Loss of functionality is often expressed by lack of motion (lock up,
stiffen, be like a stick).
% https://www.academia.edu/7973204/Towards_a_typology_of_pain_predicates_Linguistics_2012_50.3_421_466_соавторы_А.А._Бонч-Осмоловская_и_Т.И._Резникова_

Derivations locating pain or illness (\textit{-osis, -itis, -algia,}
etc), as in modern medical terminology, are quite rare.

\subsection{Temperature}
Core words for temperature may make only two distictions (``hot/warm''
vs.\ ``cold/cool''), three (``cold/cool'' vs.\ ``warm'' vs.\ ``hot''),
or four (``cold, cool, warm, hot''). Some languages may have unique
intensifiers for extreme temperatures, usually related to words for
processes or things that exemplify the temperature (such as ice or
burning).

Temperature may refer to three domains: tactile (``the plate is
hot''), ambient (``it's hot here''), and personal-feeling (``I am
hot'').  Languages like English or Italian may use a single word
(``cold,'' ``freddo'') for all three; there may a unique word for each
domain, as in East Armenian; or the domains may be split up among two
words.  Expressions for the personal-feeling domain are most likely to
be different in some way.  The ambient domain generally makes the most
distinctions, the personal-feeling the fewest.  Personal-feeling terms
may be restricted to ``uncomfortably hot'' and ``uncomfortably cold.''

Terms in the ambient domain may mark source of the heat, humidity, the
effect on breathing, windiness, etc., such as one for ``hot'' and
another for ``hot and humid.''  Some terms may be restricted to
season, such as Palula ``cool'' \textit{šidaloó}, which can refer to
pleasant coolness in summer, but is not used to describe not-too-cold
in the winter.

There may be separate tactile domain temperature terms for food and
especially water.

\subsection{Clause Connection}
% http://allegatifac.unipv.it/caterinamauri/GIACALONE_MAURIfinal.pdf
Historical sources for \textit{and:} adv.\ and prep.\ of linear
succession, ``in front, after, before, then'' (as English \textit{and}
from PIE \textit{*hanti} `in front'); focal additive particles, ``also,
too;'' paragraph linkers (``besides, moreover, and then''); comitative
markers (``with'' > ``and''); verbs meaning ``go, bring'' in narrative
contexts (Hdi \textit{là} ``to go'' > ``and''); pronominal roots (PIE
\textit{*tó} > Hittite \textit{ta} ``and'').

Historical sources for \textit{or:} distal ``that, other'' (PIE
\textit{*au-} > Lat. ``aut''); interrogative particle; free choice
verbs (Fr. ``soit... soit...''); dubitative particles ``perhaps,''
etc.; denied conditional clause ``if not, if it is not so.''

Historical sources for \textit{but:} spatial meaning of distance
(separation), closeness or opposition (OE ``be utan'' > ``but,'' ``in
stead,'' German ``sondern'' separate > ``but rather''); temporal
meaning of overlap, ``while;'' temporal meaning of continuity
``always'' (Eng ``still'' constantly > ``nonetheless''); causal and
resultative (``therefore''); comparative meaning ``more, bigger''
(Lat. \textit{magis} > It.\ ``ma'').

Semantic map of contrast: oppositive (``I bought X, but he bought Y'')
> corrective (``I'm not Xing, I'm Ying'') > counterexpectative (``I'm
tall but bad at basketball'').

\subsection{Compounding}
% ``Towards a general model of associative relations,'' https://folk.uio.no/stevepe/Pepper&Arnaud2020.pdf
There are quite a few possible relationships between the head \I{h}
and modifier \I{m} in \I{n+n} compounds.  Compounds might be formed by
1) simply cramming nouns together, as in English; 2) root + some
linking element + \I{n} (for euphonic reasons, or some other ligature
required in the language in compounds); 3) \I{n prep n} (as in French
\E{de} and Japanese \E{no}); 4) genitive; 5) \I{n} + head-marked \I{n}
(Turkish); 6) relational adjective + \I{n} (such as ``iron-\I{adj} +
road'' for ``train'' in Russian).

\begin{grammarlist}
  \item Similarity
    \begin{grammarlist}
      \item Taxonomy: an \I{m} is a kind of \I{h}: \E{oak tree;} an
        \I{h} is a kind of \I{m}: \E{bear cub}
      \item Coordination: an \I{h} that is also an \I{m}: \E{boy king}
      \item Similarity: an \I{h} that is similar to \I{m} \E{kidney bean}
    \end{grammarlist}  
  \item Containment
    \begin{grammarlist}
      \item Containment: an \I{h} that is contained in \I{m}:
        \E{orange seed;} an \I{h} that contains \I{m}: \E{seed orange}
      \item Possession: an \I{h} that is possessed by \I{m}: \E{family
        estate;} an \I{h} that possesses \I{m}: \E{career girl}
      \item Part: an \I{h} that is part of \I{m}: \E{car motor;} an
        \I{h} that \I{m} is part of: \E{motor car}
      \item Location: an \I{h} located at/near/in \I{m}: \E{house
        music;} an \I{h} that \I{m} is located at/near/in: \E{music
        hall}
      \item Time: an \I{h} that occurs at/during \I{m}: \E{summer
        job;} an \I{h} at/during which \I{m} occurs: \E{hunting season}
      \item Composition: an \I{h} that \I{m} is composed of: \E{wheat
        flour;} an \I{h} that is composed of \I{m}: \E{sugar cube}
      \item Topic: an \I{h} that is about \I{m}: \E{history book}
    \end{grammarlist}
  \item Causation (source, goal)
    \begin{grammarlist}
      \item Direction: an \I{h} whose goal is \I{m}: \E{sun worship;}
        an \I{h} that is the goal of \I{m}: \E{sales target}
      \item Source: an \I{h} that is a source of \I{m}: \E{sugar cane;}
        an \I{h} whose source is \I{m}: \E{cane sugar}
      \item Causation: an \I{h} that causes \I{m}: \E{tear gas;}
        an \I{h} that \I{m} causes: \E{sunburn}
      \item Production: an \I{h} that produces \I{m}: \E{song bird;}
        an \I{h} that \I{m} produces: \E{birdsong}
      \item Usage: an \I{h} that uses \I{m}: \E{oil lamp;}
        an \I{h} that \I{m} uses: \E{lamp oil}
      \item Function: an \I{h} that serves as \I{m}: \E{buffer state}
      \item Purpose: an \I{h} intended for \I{m}: \E{animal doctor}
    \end{grammarlist}
\end{grammarlist}

\noindent \I{h} and \I{m} elements in compounds may be metonymic
(``cottage cheese, spaghetti western, godfather, roadkill'') or
metaphorical (``milk tooth'' and the ``-father'' element in
``godfather'').


\subsection{Metaphors and Idiom}
Common source: \I{bodily manifestation for emotion}. The face often
(actions, colors), but also eyes. Other organs get up to shenanigans,
including blood.

Colors seem to often carry emotional connotations.  What a
particular color signifies will vary from culture to culture: in Thai,
a green eye or face signify anger.

Metonymy: result for whole, salient feature for whole (where whole
might be thing or event), instrument for action, effect for cause,
producer for product, both whole for part and part for whole,
perception for thing perceived (sound for event causing it).

Metonymies often chain (eye > vision > attention > desire, as in ``I
have an eye on that new computer'' vs. ``keep an eye on it''), with
idiom and metaphor applicable at all steps.  Back, buttocks \textit{>
back part > behind > after;} or \textit{behind > follow, support}.
Belly \textit{> inside part > inside > inclusive, during;} or
\textit{> pregnancy > offspring}.  Ear \textit{> hearing > attention,
(dis)regard, obedience, hearsay}.  Eye \textit{> vision > attention,
beauty}.  Head \textit{> top part > over, beginning, end}.  Mouth,
tongue \textit{> speech > word, speech act}.
% http://members.unine.ch/martin.hilpert/CMLG.pdf




\section{καὶ τὰ λοιπά}
In Kolyma Yukaghir there is a separate \I{acc} just for use on \I{1/2
sg/pl} when the subject is \I{1/2 sg/pl}, the ``pronominal
accusative.'' 

Form dependency: each element's forms may depend on choices made in
the feature above them (such as fewer tense forms in the negative, for
example).  The final three are potentially interdependent: polarity
$<$ tense, aspect, evidentiality $<$ person, reference classification
(gender, etc.) $\Leftrightarrow$ number $\Leftrightarrow$ case.

Hierarchy of diminutives and augmentatives: noun $<$ adj., verb $<$
adv., numeral, pronoun, interjection $<$ determiner.  These markers
may be distributed over several words in a sentence.

If the noun takes prefixing morphology, so will verbs.

Serial verb constructions more likely in \I{svo} languages than in
\I{sov}.  \I{sov} languages are more likely to use converbs, though a
very few have both converbs and SVCs.  Having a generalized ``and''
(\textit{i.e.,} the same for \I{np} and \I{vp}) is weakly correlated
with not having SVCs.

Pseudo-coordination (``try \uline{and} see'') is typically restricted
to a very small set of verbs, and in Germanic languages at least is
often used in aspectual constructions (``sit and ...'' in Norwegian
for progressive, ``return and...'' in Arabic for repetition).
Straight-up aspect verbs (``begin'') can also use such a
construction. ``Go'' and ``come'' with purpose may fall into this
construction.  Also used for: causative (``make/cause and''),
intention (``plan and''), reason, conditional.
% http://publish.illinois.edu/djross3/files/2014/05/Ross2014_Lisbon.pdf

Writing systems: vertical and horizontal lines, easiest to identify,
most frequent component of most symbols in a set. Obliques and
diagonals do not mix with vertical and horizontals too much (K, A, Z,
less frequent than E, H, F or W and X). Vertical symmetry (M, A, W)
more common that horizontal (K, D, E).
% http://www.shh.mpg.de/654434/legibility-emerges-spontaneously-rather-than-evolving-over-time

\subsection{Oblique Strategies}
Don't ask what it does, but how it moves.

Don't ask what it does, but what it looks like.

Don't ask what it does, but who it is important to.

Don't ask what it does, but who it hangs out with.

What is the utility of the word?

What is the default framing?  (That is, what ways to express this are
there which least offend the language speaker's Gricean maxims?)

Add something which has no history.

Verb: action vs.\ means vs.\ result.

Pick an animal and make it a central metaphor.

Pick an ancient technology and make it a central metaphor source
(stonework, ceramics, weaving, herding, sailing and knots, etc.).

Meaning comes from familiarity.

What gaps does everyone know how to fill in?  What isn't normally said
at all?

What if the arguments or referents switch animacy?

What are the hidden verb arguments?  What are the hidden
pseudo-predicate arguments?

Where is the hungry tiger?\footnote{\url{https://fiatlingua.org/2023/01/}}



\section{Lexical Functions}
\input lexfun

\section{Sound Changes}
{\small
\input changes
}

\section{Sound Systems}

\subsection{Chipaya} Onset clusters: /s ʃ/ + /p/ + (/x/); /s ʃ/ + /k q/ +
(/x xʷ χ χʷ/); /t/ + /x xʷ χ χʷ/; /tʃ l/ + /x/.  Possible codas: /x χ/
+ /p t k q l r/ + (/t/); /xʷ χʷ/ + /k q/ + (/t/); C + /t/. 

\subsection{Assiniboine} Onset clusters: /p/ + /t s ʃ tʃ/; /tk/; /k/ +
/t s ʃ tʃ m n/; /s ʃ/ + /p t k tʃ m n/; /x/ + /p t tʃ m n/; /mn/.  No
codas. 

\subsection{Oksampmin} (C)(C)V(C).  Onset choice: any C except /w j/
or labialized stop.  Onset cluster: C$_{2}$ may be /j w l x/.  Onset
/sk/ also seen.  All C may be final except prenasalized stops.

\subsection{Pech} (C)(C)V(C)(C).  Onset cluster C$_{2}$ must be /r/.
Final single: all C execpt /p t kʷ b/.  Final cluster must have /r/ as
first element.

\subsubsection{Ingessana} (C)V(C)(C).  All C in simple codas except /ʔ
k' dʒ/.  Coda cluster word finally only, C$_{1}$ is /r l n m/ and
C$_{2}$ is obstruent.

\subsubsection{Aguacatenango Tzeltal} (C)(C)V(C)(C).  Onset cluster
C$_{1}$ must be /s ʃ h/, with any consonant following.  Coda cluster
limited to /h/ + voiceless stop or affricate.

\subsubsection{Mamaindê} (C)(C)V(C)(C).  Onset cluster: /kʰ tʰ k h/ +
/w/, or /h ʔ/ + /l n j w s/, or /ʔm/.  Stops and nasals only for
simple coda, with stop or nasal + /ʔ/ for coda cluster.

\subsubsection{Qawasqar} (C)(C)(C)(C)V(C)(C)(C).  Biconsonantal onset:
/f q q' s t/ + /tʃ s t t' j w q/.  Triconsonantal onset have same
restrictions for C$_1$ (/qsq qst sqw/).  All simple codas allowed.
Complex coda: /f j l m n p q r t w s/ + /s q/.  Triconsonantal codas
include /lqs rqs qsq/.

\subsubsection{Bardi} (C)V(C)(C).  Single coda all except /p/.  Coda
cluster is /l r ɻ/ + nasal homorganic with following stop.

\subsubsection{Burushaski} (C)(C)V(C)(C).  Complex onset: /p b pʰ t d
tʰ g/ + /ɾ j/.  All C except /w j/ in simple coda.  Coda clusters:
voiceless fricative + /k/; or sonorant + /t k ʂ ɕ ts tɕ ʈʂ/.  
\end{document}
