\newcommand{\Incep}[1]{\textsc{\textbf{Incep}}(#1)}
\newcommand{\Cont}[1]{\textsc{\textbf{Cont}}(#1)}
\newcommand{\Fin}[1]{\textsc{\textbf{Fin}}(#1)}
\newcommand{\Caus}[1]{\textsc{\textbf{Caus}}(#1)}
\newcommand{\Liqu}[1]{\textsc{\textbf{Liqu}}(#1)}
\newcommand{\Perm}[1]{\textsc{\textbf{Perm}}(#1)}

\newcommand{\F}[1]{\textsc{\textbf{#1}}}
\newcommand{\fsimp}[2]{\F{#1}(\E{#2})}
\newcommand{\fmod}[3]{\F{#1}$_{#2}$(\E{#3})}

These functions are a condensed notation to represent relationships in
lexical collocation, though some will correspond to valency changes,
derivations, etc.

\subsection{Structure}
These are used for narrative and pragmatic purposes, textual cohesion,
etc., and are not simply empty.  They allow one to shift focus,
saliency, topic.  Some of this will be handled by valency tricks in
some languages.

\begin{multicols}{2}
\fmod{Func}{0}{} = \I{lvc} meaning \E{to take place} which has the
keyword lexicalized as subject: \E{the possibility exists, time flies,
the day passes by, the rain falls}.

\fmod{Func}{1}{} = \I{lvc} which connects keyword as subject with agent
as \I{do}: \E{responsibility lies (with sb), the blow comes (from sb),
support comes (from sb).} \fmod{Func}{1}{blow} = \E{comes from sb}.

\fmod{Func}{2}{} = \I{lvc} which connects keyword as subject with the
object: \F{Func}{2}{blow} = \E{falls upon sb.}

\fmod{Oper}{1}{} = ``carry out,'' \I{lvc} which connects the subject and
the action as \I{do}: \E{take a bath, vacation; have a look, bath,
shower; give sb/sth a smile, laugh, shout}.  Very common

\fmod{Oper}{2}{} = ``undergo,'' \I{lvc} which connects patient,
recipient, experiencer as subject to action as \I{do}: \E{get a
benefit, have an attack (of a disease), take advice, undergo
inspection}.

\fmod{Labor}{ij} = \I{lvc} which connects \E{i}th element as subject to
\E{j}th element as \I{do}, with keyword as secondary
object. \fmod{Labor}{12}{interrogation} = \E{to subject sb.\ to an
  interrogation;} similar, \E{treat someone with respect}.
\end{multicols}


\subsection{Fulfillment}
All these produce verbs or \I{lvc}s that satisfy ``to fulfill the
requirement of, to do with X what you are required to, X fulfills its
requirement, designed to.'' The fulfillment may not be seen as such by
someone undergoing it: the fulfullment of some disease is death.  May
be different terms fo different types of fulfillment (psychological,
physical, etc.).

\begin{multicols}{2}
\fmod{Fact}{n}{} = syntactic actant \E{n} filfulls the role indicated:
\fmod{Fact}{0}{doubt} = \E{be corroborated}.
\fmod{Fact}{1}{turn} = \E{be someone's turn}.
\fmod{Fact}{2}{ship} = \E{transport people or cargo}.
\fmod{AntiFact}{0}{accusation} = \E{is fabricated,}
\fmod{AntiFact}{1}{accusation} = \E{withdraw,}
\fmod{AntiFact}{2}{accusation} = \E{denies charges.}

\fmod{Real}{1}{} = \I{lvc} for ``act accordingly to the situation, use as
forseen:'' \E{exercise authority, use a telephone, speak a language,
  keep a promise}.

\fmod{Real}{2}{} = \I{lvc} for ``react according to the situation,''
\E{respond to an objection, satisfy a requirement, give in to
persuasion, get a joke, confirm a hypothesis.}
\fmod{AntiReal}{} = \E{fail an exam, reject advice, turn down
application}. 

\fmod{Labreal}{ij}{} = \I{lvc} corrosponding to \fsimp{Labor}{} above.
\fmod{Labreal}{12}{gallows} = \E{string someone up}.
\end{multicols}


\subsection{Verbal}
Usually from nouns.
\begin{multicols}{2}
\fsimp{Copul}{} = copula: \E{work as a teacher, server as an example,}
\fsimp{IncepCopul}{ill} = \E{fall ill}.

\fsimp{Involv}{} = involve, affect a non-participant: \E{light floods
the room, snowstorm catches/hits}.
  
\fsimp{Manif}{} = manifest, become apparent: \E{joy bursts, scorn
drips}.

\fsimp{Prox}{} = be about to, be on the verge of: \E{on the edge of
despair, on the brink of disaster, verge of tears, thunderstorm
brews}.  Usually as \fmod{ProxOper}{1}{despair}.

\fsimp{Prepar}{} = prepare X for, get X ready for normal use.  Usually
as \fmod{PerparFact}{0}{car} = \E{fill up the car}.

\fsimp{Obstr}{} = function with difficulty: \E{eyes blur, economy
stagnates, short of breath}. \fsimp{CausObstr}{} common: \E{gun
jams, rope tangles, traffic snarls}.

\fsimp{Son}{} = emit characteristic sound: \E{whip cracks, bell chimes,
cane swish, leaf rustles}.

\fsimp{Stop}{} = stop functioning: \E{lose one's breath, voice breaks,
heart stops (or breaks)}.

\fsimp{Excess}{} = function in an abnormally excessive way: \E{heart
has palpitations, sweat rolls down, theeth grind}.

\fsimp{Sympt}{} = represents bodily reaction to X.  Joined in complex
relationships with the rest:
\fsimp{Obstr}{speech} + \fsimp{Sympt}{anger} = \E{sputters with anger;} 
\fsimp{Obstr}{breath} + \fsimp{Sympt}{anger} = \E{chokes with anger}.
\end{multicols}


\subsection{Nominal}
~
\begin{multicols}{2}
\fmod{S}{n}{} = the \E{n}th participant (agent noun, object noun,
etc.). 
\fmod{S}{1}{teach} = \E{teacher,}
\fmod{S}{2}{teach} = \E{student,}
\fmod{S}{3}{teach} = \E{subject matter.}
More interesting when modified, e.g., \fmod{(AntiBon)S}{1}{}.

\fsimp{Cap}{} = the head of: \E{pope, captain, emperor,}
\fsimp{Cap}{university} = \E{president}.

\fsimp{Equip}{} = staff, crew of: \E{crew, company, personnel}.

\fsimp{Mult}{} = collection of: \E{bouquet, group}.

\fsimp{Pel}{} = covering: \fsimp{Pel}{bean} = \E{pod,}
\fsimp{Pel}{tree} = \E{bark,}
\fsimp{Pel}{book} = \E{cover; binding.}
\fsimp{LiquPel}{bean} = \E{to shell,} \fsimp{LiquPel}{rabbit} = \E{to skin}.

\fsimp{Fas}{} = ``face,'' front: \E{front of house, bow or prow of
ship, nose of plane.}  \fsimp{AntiFas}{} = \E{ship stern, tail of
plane}. 
\end{multicols}

\subsection{Adjectival}
~
\begin{multicols}{2}
\fmod{A}{n}{} = determining property of \E{n}th participant from the
viewpoint of its role in the situation; quite like participles with verbs.
\fmod{A}{0}{brother} = \E{fraternal,}
\fmod{A}{0}{city} = \E{urban,}
\fmod{A}{1}{delight} = \E{delightful,}
\fmod{A}{1}{anger} = \E{in anger, angry,}
\fmod{A}{2}{shoot} = \E{under fire}.
\fmod{A}{2}{analyze} = \E{under analysis}.
Often more useful modified.

\fmod{Able}{n}{} = can easily, prone to:
  \fmod{Able}{1}{cry} = \E{tearful,}
  \fmod{Able}{2}{trust} = \E{trustworthy}.

\fmod{Qual}{i}{} = predisposed, of \E{i}th probable argument: 
  \fmod{Qual}{1}{cry} = \E{sad,}
  \fmod{Qual}{1}{laugh} = \E{cheerful,}
  \fmod{Qual}{2}{laugh} = \E{awkward, absurd}.
\end{multicols}


\subsection{Evaluation}
May be combined with \F{Anti}: \fsimp{Magn}{temperature} = \E{high},
\fsimp{AntiMagn}{temperature} = \E{low}.  May be quite different
depending on word class, and take multiple forms: \fsimp{Magn}{smoker}
= \E{heavy, chain-smoker;} \fsimp{Magn}{to smoke} = \E{like a
chimney}. 

\begin{multicols}{2}
\fsimp{Bon}{} = good, generally held praise: \E{neatly cut, heroic
  struggle, fruitful analysis}.

\fsimp{Centr}{} = center, culmination: \E{height of the crisis, summit
of glory, prime of life}.

\fsimp{Degrad}{} = degraded, lowered: \E{discipline decays, house becomes
dilapidated, patience wears thin, temper frays, teeth decay}.

\fsimp{Magn}{} = immensely, very: \E{shave close/clean, condemn
strongly, infinite patience}.  Might be quantitative or temporal
(speed). 

\fsimp{Ver}{} = real, genuine, as it should be, meeting intended
requirements: \E{genuine surprise, walk steadily, loyal citizen,
legitimate demand, precise instrument, well-deserved punishment,
restful sleep}. 

\fsimp{AntiVer}{} might have ``too much'' or ``too little'' options.
\end{multicols}


\subsection{Other}
~
\begin{multicols}{2}
\fsimp{Result}{} = the expected result of;
\fsimp{Result}{buy} = \E{own,}
\fsimp{Result}{to have learnt} =\E{know}.

\fmod{Conv}{ijk}{} = converse, reorders arguments.
\fmod{Conv}{21}{include} = \E{belong,}
\fmod{Conv}{21}{precede} = \E{follow}.
  
\fsimp{Figur}{} = figurative, standard received metaphors: \E{curtain
of rain, pangs of remorse, flames of passion}.
\end{multicols}


\subsection{Modification}
These are only used in combination with others.
\fmod{IncepFunc}{1}{anger} = \E{anger rises}.
\fmod{ContOper}{1}{power} = \E{retain one's power}.
\fmod{PermFunc}{0}{aggression} = \E{condone aggression}

\begin{multicols}{2}
\Incep{begin, start} (= \fsimp{AntiFin}{})

\Cont{continue, maintain, retain}

\Fin{cease, stop}

\Caus{causative}

\Liqu{liquidate, stop, divert} (= \fsimp{AntiCaus}{})

\Perm{permit, allow, condone}

\F{Plus}(more)

\F{Minus}(less)

\F{Anti}(negates)
\end{multicols}

\noindent Apparently common blends:
\begin{multicols}{2}
\F{AntiMagn}

\fmod{CausFunc}{0}{} = \E{find an answer, conduct a campaign, produce
  an effect}.  Fairly common.

\fmod{CausFunc}{1}{} = \E{open the way, cause damage, give an
  answer}.  Fairly common.  Can be modified:
\fmod{CausPlusFunc}{1}{risk} = \E{increase, raise,}
\fmod{CausMinusFunc}{1}{consumption} = \E{reduce}

\fmod{IncepOper}{1}{} = \E{take an attitude, start a session, obtain a
  position}.

\fmod{ContOper}{1}{} = \E{keep silence, follow an example, keep one's
  balance, lead a busy life}.
\end{multicols}

\noindent A fuller example: \\
\fmod{IncepOper}{1}{habit} = \E{acquire, form, take to}.\\
\fmod{FinOper}{1}{habit} = \E{drop, get out/rid of}.\\
\fmod{LiquOper}{1}{habit} = \E{break, wean from}.\\
\fmod{Liqu$_{1}$Oper}{1}{habit} = \E{kick, shake off}.\\
\fmod{CausFunc}{1}{habit} = \E{instill into, inculcate}.

\bigskip
These may also produce simultaneous functions,
\F{[Magn + Oper$_{1}$]}(doubt) = \E{be plagued by doubt}.
\F{[Ver + Oper$_{1}$]}(health) = \E{have a clean bill of health}.
